---
title: Times 60-mock vs 0 contrast
---

```{r}
#| include: false

library("DESeq2")
library("clusterProfiler")
library("org.Bthuringiensis.eg.db")
library("dplyr")
library("tidyr")
library("ggplot2")
library("plotly")
library("DT")
```

```{r}
#| include: false

source("rnaseq_functions.R")
```

```{r}
#| include: false

counts <- readRDS(file = "lytic_cycle_induction_gil01_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_gil01_metadata.rds")
all_genes <- readRDS(file = "lytic_cycle_induction_gil01_all_genes.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time)

dds <- DESeq(dds)
```

In order to perform the times 60-mock vs 0 contrast, we established a log₂ fold-change (LFC) threshold of ±0.585. This value reflects an increase or decrease in expression of 50%, depending on whether it is positive or negative, respectively. The default value is 0, but when performing the other contrasts we obtained too many differentially expressed (DE) genes. This violates one of DESeq2's key assumptions, which is that most genes are not differentially expressed.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: contrast
#| code-fold: true

contrast <- results(dds, contrast = c("Time", "60_mock", 0), lfcThreshold = log2(1.5))
contrast_df <- as.data.frame(contrast)

contrast_df <- contrast_df %>%
  { x <- .; x$Gene_ID <- rownames(x); x } %>%
  left_join(
    annotation %>%
      filter(Gene_ID %in% rownames(contrast_df)) %>%
      dplyr::select(Gene_ID, Gene_symbol, Description, COG_category, Association, Type, Subtype, Source),
    by = "Gene_ID"
  ) %>%
  rename(
    Mean_normalized_counts = baseMean,
    LFC = log2FoldChange,
    Adjusted_p_value = padj
  ) %>%
  dplyr::select(
    Gene_ID,
    Gene_symbol,
    Description,
    COG_category,
    Association,
    Type,
    Subtype,
    Source,
    Mean_normalized_counts,
    LFC,
    Adjusted_p_value
  )
```

As can be seen in the `summary()` results, 248 upregulated and 212 downregulated genes were obtained. In addition, 293 genes were discarded due to low counts, probably because they are expressed in inducted samples or in the absence of GIL01. Below is an interactive table where all DE genes can be viewed individually.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: general_results
#| code-fold: true

summary(contrast, alpha = 0.05)

datatable(contrast_df[!is.na(contrast_df$Adjusted_p_value) & contrast_df$Adjusted_p_value < 0.05, ],
          colnames = c("Gene ID", "Gene symbol", "Description", "COG category", "Association",
          "Type", "Subtype", "Source", "Mean normalized counts", "LFC", "Adjusted p value"), rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2, rightColumns = 2),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = c("Mean_normalized_counts", "LFC"), digits = 2) %>%
  formatSignif(columns = "Adjusted_p_value", digits = 3)
```

# MA plot

A MA plot visualizes the relationship between the mean expression and the LFC for each gene. It helps to identify genes that are differentially expressed across conditions, highlighting those with large fold changes relative to their expression level.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ma_plot
#| code-fold: true
#| message: false

plotMA(contrast, alpha = 0.05, xlab = "Mean of normalized counts", ylab = "LFC")
```

# DE genes

DE genes are defined as those with an adjusted p-value below 0.05. Among them, upregulated genes are extracted as those with a LFC > 0.58, while downregulated genes are defined as those with an LFC < -0.58.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: de_genes
#| code-fold: true

up_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange > log2(1.5) & !is.na(padj)))
down_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange < -log2(1.5) & !is.na(padj)))
```

# Volcano plot

A volcano plot displays each gene according to its LFC on the x-axis and the statistical significance (−log₁₀ adjusted p-value) on the y-axis. In this contrast, we found a large number of upregulated and downregulated genes on the chromosome, but on pBtic235 we only found an upregulated gene (GBJ002-DNT_251) and some downregulated genes which encode primarily for tRNA and phage and hypothetical proteins. As we could see in the analog contrast for GBJ002, GBJ002-DNT_251 is the most upregulated pBtic235 gene, so it appears to be more abundant over time regardless of MMC induction and presence of GIL01. Furthermore, as expected, no GIL01 gene is DE, even though GIL01-DNT_01 has a LFC \> 1 (adjusted p-value = 0.582). In the absence of further data, it could be interpreted that this gene is involved in the maintenance of lysogeny and/or has an anti-sense interference mechanism of action, as explained by [Pavlin *et al*. (2022)](https://doi.org/10.1038/s42003-022-04238-3).

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: volcano_plot
#| code-fold: true

volcano_plot(contrast_df, limits = c(-8, 8))
```

# Enrichment analysis

In this section, we present the results of enrichment analyses based on the COG (Clusters of Orthologous Groups), KEGG (Kyoto Encyclopedia of Genes and Genomes), GO (Gene Ontology) and EC (Enzyme Commission) categories for both upregulated and downregulated genes.

## COG

According to COG categories, we only found positive enrichments in energy production and conversion (**C**), carbohydrate transport and metabolism (**G**), amino acid transport and metabolism (**E**), nucleotide transport and metabolism (**F**) and lipid transport and metabolism (**I**) by 28, 29, 33, 14 and 15 genes, respectively. As we could see in the GBJ002 analogous contrast, in this case we can observe more enrichments, suggesting that, even in the absence of MMC induction, the presence of GIL01 affects GBJ002 metabolism.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog
#| code-fold: true

cog_up <- cog_enrichment(up_genes, all_genes)
cog_down <- cog_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog_results
#| code-fold: true

datatable(cog_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(cog_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

## KEGG

According to KEGG categories, we found positive and negative enrichments. In the case of the former, they were all related to metabolism: amino acids, carbohydrates, biosynthesis of secondary metabolites, etc. This was to be expected, since the culture was growing and the metabolism was active. In the case of the latter, it resulted in depletion of other metabolic pathways, such as biosynthesis of other amino acids and nitrogen metabolism.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg
#| message: false
#| code-fold: true

kegg_up <- kegg_enrichment(up_genes, all_genes)
kegg_down <- kegg_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg_results
#| code-fold: true

datatable(kegg_up, rownames = FALSE, caption = "Positive enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(kegg_down, rownames = FALSE, caption = "Negative enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## GO

The Gene Ontology (GO) is organized into three categories: **Biological Process** (BP), which describes the biological objectives to which gene products contribute; **Molecular Function** (MF), which refers to the elemental activities of gene products at the molecular level; and **Cellular Component** (CC), which specifies the locations within the cell where gene products are active.

-   According to **BP**, positive enrichment was mainly related to amino acids, carbohydrates and nucleic acids metabolism, while negative enrichment was found in organic acids, oxoacids and other amino acids processes.

-   Regarding **MF**, we found a positive enrichment in oxidoreductase activity.

-   Regarding **CC**, we found positive enrichments in oxidoreductase and catalytic complexes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go
#| code-fold: true
#| message: false

go_bp_up <- go_enrichment(up_genes, all_genes, "BP")
go_bp_down <- go_enrichment(down_genes, all_genes, "BP")

go_mf_up <- go_enrichment(up_genes, all_genes, "MF")
go_mf_down <- go_enrichment(down_genes, all_genes, "MF")

go_cc_up <- go_enrichment(up_genes, all_genes, "CC")
go_cc_down <- go_enrichment(down_genes, all_genes, "CC")
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go_results
#| code-fold: true

datatable(go_bp_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_bp_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## EC

The EC classification establishes up to four levels to define enzymes. In this section, the first three were analyzed, although statistical significance was only achieved in the first (**1**) and second (**1.4**) levels due to upregulated oxidoreductases and GBJ002_01383 (leucine dehydrogenase), GBJ002_01453 (aminomethyl-transferring glycine dehydrogenase subunit 2), GBJ002_01454 (aminomethyl-transferring glycine dehydrogenase subunit 1), GBJ002_01836 (alanine dehydrogenase) and GBJ002_04256 (NAD-specific glutamate dehydrogenase) genes, respectively.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec
#| code-fold: true

ec_1_up <- ec_enrichment(up_genes, all_genes, 1)
ec_1_down <- ec_enrichment(down_genes, all_genes, 1)

ec_2_up <- ec_enrichment(up_genes, all_genes, 2)
ec_2_down <- ec_enrichment(down_genes, all_genes, 2)

ec_3_up <- ec_enrichment(up_genes, all_genes, 3)
ec_3_down <- ec_enrichment(down_genes, all_genes, 3)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec_results
#| code-fold: true

datatable(ec_1_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_1_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

# Enrichment summary

This table contains a summary of the enrichments for each category. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: enrichment_summary
#| code-fold: true

es_up <- enrichment_summary(up_genes, categories = list(cog_up, kegg_up, go_bp_up, go_mf_up, go_cc_up, ec_1_up, ec_2_up, ec_3_up))
es_down <- enrichment_summary(down_genes, categories = list(cog_down, kegg_down, go_bp_down, go_mf_down, go_cc_down, ec_1_down, ec_2_down, ec_3_down))
es_de <- rbind(es_up, es_down)
rownames(es_de) <- c("UP", "DOWN")

datatable(es_de, colnames = c("DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2),
            scrollX = TRUE,
            scroller = TRUE,
            deferRender = TRUE))
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

create_summary(contrast_name = "Times 60-mock vs 0", rds_name = "gil01_t60-mock_t0")
```
