---
title: Times 60 vs 60-mock contrast
---

```{r}
#| include: false

library("DESeq2")
library("clusterProfiler")
library("org.Bthuringiensis.eg.db")
library("dplyr")
library("tidyr")
library("ggplot2")
library("plotly")
library("DT")
```

```{r}
#| include: false

source("rnaseq_functions.R")
```

```{r}
#| include: false

counts <- readRDS(file = "lytic_cycle_induction_gil01_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_gil01_metadata.rds")
all_genes <- readRDS(file = "lytic_cycle_induction_gil01_all_genes.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time + MMC)

dds <- DESeq(dds)
```

In order to perform the times 60 vs 60-mock contrast, we established a log₂ fold-change (LFC) threshold of ±0.585. This value reflects an increase or decrease in expression of 50%, depending on whether it is positive or negative, respectively. The default value is 0, but when performing the other contrasts we obtained too many differentially expressed (DE) genes. This violates one of DESeq2's key assumptions, which is that most genes are not differentially expressed.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: contrast
#| code-fold: true

contrast <- results(dds, name = "MMC_TRUE_vs_FALSE", lfcThreshold = log2(1.5))
contrast_df <- as.data.frame(contrast)

contrast_df <- contrast_df %>%
  { x <- .; x$Gene_ID <- rownames(x); x } %>%
  left_join(
    annotation %>%
      filter(Gene_ID %in% rownames(contrast_df)) %>%
      dplyr::select(Gene_ID, Gene_symbol, Description, COG_category, Association, Type, Subtype, Source),
    by = "Gene_ID"
  ) %>%
  rename(
    Mean_normalized_counts = baseMean,
    LFC = log2FoldChange,
    Adjusted_p_value = padj
  ) %>%
  dplyr::select(
    Gene_ID,
    Gene_symbol,
    Description,
    COG_category,
    Association,
    Type,
    Subtype,
    Source,
    Mean_normalized_counts,
    LFC,
    Adjusted_p_value
  )
```

As can be seen in the `summary()` results, 751 upregulated and 600 downregulated genes were obtained. Below is an interactive table where all DE genes can be viewed individually.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: general_results
#| code-fold: true

summary(contrast, alpha = 0.05)

datatable(contrast_df[!is.na(contrast_df$Adjusted_p_value) & contrast_df$Adjusted_p_value < 0.05, ],
          colnames = c("Gene ID", "Gene symbol", "Description", "COG category", "Association",
          "Type", "Subtype", "Source", "Mean normalized counts", "LFC", "Adjusted p value"), rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2, rightColumns = 2),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = c("Mean_normalized_counts", "LFC"), digits = 2) %>%
  formatSignif(columns = "Adjusted_p_value", digits = 3)
```

# MA plot

A MA plot visualizes the relationship between the mean expression and the LFC for each gene. It helps to identify genes that are differentially expressed across conditions, highlighting those with large fold changes relative to their expression level.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ma_plot
#| code-fold: true
#| message: false

plotMA(contrast, alpha = 0.05, xlab = "Mean of normalized counts", ylab = "LFC")
```

# DE genes

DE genes are defined as those with an adjusted p-value below 0.05. Among them, upregulated genes are extracted as those with a LFC > 0.58, while downregulated genes are defined as those with an LFC < -0.58.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: de_genes
#| code-fold: true

up_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange > log2(1.5) & !is.na(padj)))
down_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange < -log2(1.5) & !is.na(padj)))
```

# Volcano plot

A volcano plot displays each gene according to its LFC on the x-axis and the statistical significance (−log₁₀ adjusted p-value) on the y-axis. In this contrast, we found a large number of upregulated and downregulated genes on both the chromosome and pBtic235. Regarding the prophage region, it contains 9 of the 10 most overexpressed genes of the chromosome, the remaining one being GBJ002-DNT_226. Regarding GIL01, almost all genes were upregulated and with LFC between 5.17 and 10.58, being GIL01-DNT_01 the only not DE gene. In the plot, if we select only upregulated GIL01 genes, we can distinguish four zones: GIL01_01 to GIL01_07 genes, with LFC \> 5 and adjusted p-values \< 10^-36^; GIL01_09 to GIL01_30 genes (and GIL01_35), with LFC \> 8 and adjusted p-values \< 10^-37^; GIL01_36, with LFC = 10.58 and adjusted p-value ≈ 10^-14^; and GIL01_08, GIL01_33 and GIL01_34 genes, with LFC \< 7 and adjusted p-values \> 10^-16^. The first region corresponds mainly to the genes of the P1-P2 tandem promoters, while the second corresponds to those of the P3 promoter. The third region corresponds to a CDS predicted by Pharokka that may not be real, since its mean normalized counts are less than 70. Something similar occurs with the genes in the fourth region, which are poorly expressed. On the one hand, GIL01_08 has a mean normalized count of approximately 200. On the other hand, the genes expressed from the negative strand, GIL01_33 and GIL01_34, are found. In any case, the important thing about these areas is to highlight that the genes expressed from the P3 promoter are much more upregulated than those expressed from the P1-P2 tandem promoters.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: volcano_plot
#| code-fold: true

volcano_plot(contrast_df, limits = c(-11, 11))
```

# Enrichment analysis

In this section, we present the results of enrichment analyses based on the COG (Clusters of Orthologous Groups), KEGG (Kyoto Encyclopedia of Genes and Genomes), GO (Gene Ontology) and EC (Enzyme Commission) categories for both upregulated and downregulated genes.

## COG

According to COG categories, we found positive enrichment in replication, recombination and repair (**L**), inorganic ion transport and metabolism (**P**) and carbohydrate transport and metabolism (**G**) and negative enrichment in transcription (**K**) and posttranslational modification, protein turnover and chaperones (**O**).

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog
#| code-fold: true

cog_up <- cog_enrichment(up_genes, all_genes)
cog_down <- cog_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog_results
#| code-fold: true

datatable(cog_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(cog_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

## KEGG

According to KEGG categories, we found positive and negative enrichments. In the case of the former, they were mainly related to replication and repair pathways, but also to translation and energy metabolism. In the case of the latter, it resulted only in depletion of quorum sensing pathways due to 20 downregulated genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg
#| message: false
#| code-fold: true

kegg_up <- kegg_enrichment(up_genes, all_genes)
kegg_down <- kegg_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg_results
#| code-fold: true

datatable(kegg_up, rownames = FALSE, caption = "Positive enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(kegg_down, rownames = FALSE, caption = "Negative enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## GO

The Gene Ontology (GO) is organized into three categories: **Biological Process** (BP), which describes the biological objectives to which gene products contribute; **Molecular Function** (MF), which refers to the elemental activities of gene products at the molecular level; and **Cellular Component** (CC), which specifies the locations within the cell where gene products are active.

-   According to **BP**, we observed positive enrichment in processes related to cellular response to stimuli, DNA processes and ions transport, among others. Negative enrichment was found in metabolic processes associated with amino acids and other compounds, such as organic acids and small molecules.

-   Regarding **MF**, we only found positive enrichment in processes related to transporter, DNA-associated and RNA binding activities.

-   Regarding **CC**, positive enrichments were related to ribosomes, intracelullar organelles and protein-containing complexes, among other components.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go
#| code-fold: true
#| message: false

go_bp_up <- go_enrichment(up_genes, all_genes, "BP")
go_bp_down <- go_enrichment(down_genes, all_genes, "BP")

go_mf_up <- go_enrichment(up_genes, all_genes, "MF")
go_mf_down <- go_enrichment(down_genes, all_genes, "MF")

go_cc_up <- go_enrichment(up_genes, all_genes, "CC")
go_cc_down <- go_enrichment(down_genes, all_genes, "CC")
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go_results
#| code-fold: true

datatable(go_bp_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_bp_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## EC

The EC classification establishes up to four levels to define enzymes. In this section, the first three were analyzed, achieving statistical significance in all of them for positive enrichment. In the first level, oxidoreductases (**1**) and translocases (**7**) numbers were enriched by 36 and 14 DE genes, respectively. In the second, **5.6** level was enriched by 9 genes. Lastly, in the third, **5.6.2** level was enriched by 9 out of 16 genes, corresponding to GBJ002_00759 (DNA topoisomerase IV subunit A), GBJ002_00760 (DNA topoisomerase IV subunit B), GBJ002_00997 (replicative DNA helicase), GBJ002_01078 (primosomal protein N'), GBJ002_02711 (replicative DNA helicase), GBJ002_02736 (DNA topoisomerase (ATP-hydrolyzing) subunit B), GBJ002_02737 (DNA gyrase subunit A), GBJ002_03710 (replicative DNA helicase) and GBJ002_03905 (helicase-exonuclease AddAB subunit AddA) genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec
#| code-fold: true

ec_1_up <- ec_enrichment(up_genes, all_genes, 1)
ec_1_down <- ec_enrichment(down_genes, all_genes, 1)

ec_2_up <- ec_enrichment(up_genes, all_genes, 2)
ec_2_down <- ec_enrichment(down_genes, all_genes, 2)

ec_3_up <- ec_enrichment(up_genes, all_genes, 3)
ec_3_down <- ec_enrichment(down_genes, all_genes, 3)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec_results
#| code-fold: true

datatable(ec_1_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_1_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

# Enrichment summary

This table contains a summary of the enrichments for each category. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: enrichment_summary
#| code-fold: true

es_up <- enrichment_summary(up_genes, categories = list(cog_up, kegg_up, go_bp_up, go_mf_up, go_cc_up, ec_1_up, ec_2_up, ec_3_up))
es_down <- enrichment_summary(down_genes, categories = list(cog_down, kegg_down, go_bp_down, go_mf_down, go_cc_down, ec_1_down, ec_2_down, ec_3_down))
es_de <- rbind(es_up, es_down)
rownames(es_de) <- c("UP", "DOWN")

datatable(es_de, colnames = c("DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2),
            scrollX = TRUE,
            scroller = TRUE,
            deferRender = TRUE))
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

create_summary(contrast_name = "Times 60 vs 60-mock", rds_name = "gil01_t60_t60-mock")
```
