---
title: GIL16 vs GIL01 contrast
---

```{r}
#| include: false

library("DESeq2")
library("clusterProfiler")
library("org.Bthuringiensis.eg.db")
library("dplyr")
library("tidyr")
library("ggplot2")
library("plotly")
library("DT")
```

```{r}
#| include: false

source("rnaseq_functions.R")
```

```{r}
#| include: false

counts <- readRDS(file = "lytic_cycle_induction_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_metadata.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")

counts[] <- lapply(counts, function(x) if(is.numeric(x)) ifelse(x %% 1 == 0.5, ceiling(x), round(x)) else x)

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time + Strain + Time:Strain)
dds <- DESeq(dds)
dds <- dds[!grepl("^GIL", rownames(dds)), ]

all_genes <- rownames(dds)
```

To analyze differential expression across strains and time points, we performed all statistical modeling within a single DESeq2 dataset rather than creating separate DESeq2 objects for each time point. Because the experimental design includes strong temporal dynamics associated with phage induction, library composition changes substantially across conditions. Analyzing all samples together allows DESeq2 to estimate size factors and dispersion parameters using the full dataset, thereby borrowing statistical strength across conditions and ensuring a consistent normalization framework. Importantly, replicate effects (treated as a batch factor in the model) are captured globally; splitting the dataset by time would force DESeq2 to re‑estimate normalization and dispersion independently for each subset, disrupting the shared technical structure of the experiment and potentially introducing artificial differences. Using a unified DESeq2 object therefore provides a coherent, statistically robust foundation that preserves biological comparability, controls technical variation and enables accurate strain‑level contrasts at each time point.

In order to perform the GIL16 vs GIL01 contrast at 60 minutes, we established a log₂ fold-change (LFC) of ±0.585. This value reflects an increase or decrease in expression of 50%, depending on whether it is positive or negative, respectively. The default value is 0, but when performing the other contrasts we obtained too many differentially expressed (DE) genes. This violates one of DESeq2's key assumptions, which is that most genes are not differentially expressed.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: contrast
#| code-fold: true

contrast <- results(dds, contrast = list(c("Strain_GIL16_vs_GBJ002", "Time60.StrainGIL16"), c("Strain_GIL01_vs_GBJ002", "Time60.StrainGIL01")), lfcThreshold = log2(1.5))
contrast_df <- as.data.frame(contrast)

contrast_df <- contrast_df %>%
  { x <- .; x$Gene_ID <- rownames(x); x } %>%
  left_join(
    annotation %>%
      filter(Gene_ID %in% rownames(contrast_df)) %>%
      dplyr::select(Gene_ID, Gene_symbol, Description, COG_category, Association, Type, Subtype, Source),
    by = "Gene_ID"
  ) %>%
  rename(
    Mean_normalized_counts = baseMean,
    LFC = log2FoldChange,
    Adjusted_p_value = padj
  ) %>%
  dplyr::select(
    Gene_ID,
    Gene_symbol,
    Description,
    COG_category,
    Association,
    Type,
    Subtype,
    Source,
    Mean_normalized_counts,
    LFC,
    Adjusted_p_value
  )
```

As can be seen in the `summary()` results, only 2 upregulated genes were obtained. Below is an interactive table where all DE genes can be viewed individually.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: general_results
#| code-fold: true

summary(contrast, alpha = 0.05)

datatable(contrast_df[!is.na(contrast_df$Adjusted_p_value) & contrast_df$Adjusted_p_value < 0.05, ],
          colnames = c("Gene ID", "Gene symbol", "Description", "COG category", "Association",
          "Type", "Subtype", "Source", "Mean normalized counts", "LFC", "Adjusted p value"), rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2, rightColumns = 2),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = c("Mean_normalized_counts", "LFC"), digits = 2) %>%
  formatSignif(columns = "Adjusted_p_value", digits = 3)
```

# MA plot

A MA plot visualizes the relationship between the mean expression and the LFC for each gene. It helps to identify genes that are differentially expressed across conditions, highlighting those with large fold changes relative to their expression level.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ma_plot
#| code-fold: true
#| message: false

plotMA(contrast, alpha = 0.05, xlab = "Mean of normalized counts", ylab = "LFC")
```

# DE genes

DE genes are defined as those with an adjusted p-value below 0.05. Among them, upregulated genes are extracted as those with a LFC > 0.58, while downregulated genes are defined as those with an LFC < -0.58.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: de_genes
#| code-fold: true

up_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange > log2(1.5) & !is.na(padj)))
down_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange < -log2(1.5) & !is.na(padj)))
```

# Volcano plot

A volcano plot displays each gene according to its LFC on the x-axis and the statistical significance (−log₁₀ adjusted p-value) on the y-axis. Only the isocitrate lyase (GBJ002_03895) and DUF4153 domain-containing protein (GBJ002_04612) genes were upregulated, i.e., more expressed in the GIL16 samples than in the GIL01 samples.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: volcano_plot
#| code-fold: true

volcano_plot(contrast_df, limits = c(-5, 5))
```

# Enrichment analysis

In this section, we present the results of enrichment analyses based on the COG (Clusters of Orthologous Groups), KEGG (Kyoto Encyclopedia of Genes and Genomes), GO (Gene Ontology) and EC (Enzyme Commission) categories for both upregulated and downregulated genes.

## COG

According to COG categories, no enrichments were found.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog
#| code-fold: true

cog_up <- cog_enrichment(up_genes, all_genes)
cog_down <- cog_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog_results
#| code-fold: true

datatable(cog_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(cog_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

## KEGG

According to KEGG categories, no enrichments were found.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg
#| message: false
#| code-fold: true

kegg_up <- kegg_enrichment(up_genes, all_genes)
kegg_down <- kegg_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg_results
#| code-fold: true

datatable(kegg_up, rownames = FALSE, caption = "Positive enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(kegg_down, rownames = FALSE, caption = "Negative enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## GO

The Gene Ontology (GO) is organized into three categories: **Biological Process** (BP), which describes the biological objectives to which gene products contribute; **Molecular Function** (MF), which refers to the elemental activities of gene products at the molecular level; and **Cellular Component** (CC), which specifies the locations within the cell where gene products are active. In any case, no enrichments were found.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go
#| code-fold: true
#| message: false

go_bp_up <- go_enrichment(up_genes, all_genes, "BP")
go_bp_down <- go_enrichment(down_genes, all_genes, "BP")

go_mf_up <- go_enrichment(up_genes, all_genes, "MF")
go_mf_down <- go_enrichment(down_genes, all_genes, "MF")

go_cc_up <- go_enrichment(up_genes, all_genes, "CC")
go_cc_down <- go_enrichment(down_genes, all_genes, "CC")
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go_results
#| code-fold: true

datatable(go_bp_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_bp_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## EC

The EC classification establishes up to four levels to define enzymes. In this section, the first three were analyzed, although no statistically significant results were obtained in any case.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec
#| code-fold: true

ec_1_up <- ec_enrichment(up_genes, all_genes, 1)
ec_1_down <- ec_enrichment(down_genes, all_genes, 1)

ec_2_up <- ec_enrichment(up_genes, all_genes, 2)
ec_2_down <- ec_enrichment(down_genes, all_genes, 2)

ec_3_up <- ec_enrichment(up_genes, all_genes, 3)
ec_3_down <- ec_enrichment(down_genes, all_genes, 3)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec_results
#| code-fold: true

datatable(ec_1_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_1_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

# Enrichment summary

This table contains a summary of the enrichments for each category. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: enrichment_summary
#| code-fold: true

es_up <- enrichment_summary(up_genes, categories = list(cog_up, kegg_up, go_bp_up, go_mf_up, go_cc_up, ec_1_up, ec_2_up, ec_3_up))
es_down <- enrichment_summary(down_genes, categories = list(cog_down, kegg_down, go_bp_down, go_mf_down, go_cc_down, ec_1_down, ec_2_down, ec_3_down))
es_de <- rbind(es_up, es_down)
rownames(es_de) <- c("UP", "DOWN")

datatable(es_de, colnames = c("DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2),
            scrollX = TRUE,
            scroller = TRUE,
            deferRender = TRUE))
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

create_summary(contrast_name = "GIL16 vs GIL01", rds_name = "gil16_gil01_t60")
```
