---
title: Time points 30 vs 0 contrast
---

```{r}
#| include: false

library("DESeq2")
library("clusterProfiler")
library("org.Bthuringiensis.eg.db")
library("dplyr")
library("tidyr")
library("ggplot2")
library("plotly")
library("DT")
```

```{r}
#| include: false

source("tectivirnaseq_functions.R")
```

```{r}
#| include: false

counts <- readRDS(file = "lytic_cycle_induction_gil16_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_gil16_metadata.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time + MMC)
dds <- DESeq(dds)
all_genes <- rownames(dds)
```

In order to perform the time 30 vs 0 contrast, we established a log₂ fold-change (LFC) threshold of ±0.585. This value reflects an increase or decrease in expression of 50%, depending on whether it is positive or negative, respectively. The default value is 0, but when performing the other contrasts we obtained too many differentially expressed (DE) genes. This violates one of DESeq2's key assumptions, which is that most genes are not differentially expressed.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: contrast
#| code-fold: true

contrast <- results(dds, contrast = list(c("Time_30_vs_0", "MMC_TRUE_vs_FALSE")), lfcThreshold = log2(1.5))
contrast_df <- as.data.frame(contrast)

contrast_df <- contrast_df %>%
  { x <- .; x$Gene_ID <- rownames(x); x } %>%
  left_join(
    annotation %>%
      filter(Gene_ID %in% rownames(contrast_df)) %>%
      dplyr::select(Gene_ID, Gene_symbol, Description, Classification, Association, Type, Subtype, Source),
    by = "Gene_ID"
  ) %>%
  rename(
    Mean_normalized_counts = baseMean,
    LFC = log2FoldChange,
    Adjusted_p_value = padj
  ) %>%
  dplyr::select(
    Gene_ID,
    Gene_symbol,
    Description,
    Classification,
    Association,
    Type,
    Subtype,
    Source,
    Mean_normalized_counts,
    LFC,
    Adjusted_p_value
  )
```

As can be seen in the `summary()` results, 277 upregulated and 312 downregulated genes were obtained. In addition, 293 genes were discarded due to low counts, probably because they are expressed in latter times or in the absence of GIL16. Below is an interactive table where all DE genes can be viewed individually.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: general_results
#| code-fold: true

summary(contrast, alpha = 0.05)

datatable(contrast_df[!is.na(contrast_df$Adjusted_p_value) & contrast_df$Adjusted_p_value < 0.05, ],
          colnames = c("Gene ID", "Gene symbol", "Description", "COG category / Homologue", "Association",
          "Type", "Subtype", "Source", "Mean normalized counts", "LFC", "Adjusted p value"), rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2, rightColumns = 2),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = c("Mean_normalized_counts", "LFC"), digits = 2) %>%
  formatSignif(columns = "Adjusted_p_value", digits = 3)
```

# MA plot

A MA plot visualizes the relationship between the mean expression and the LFC for each gene. It helps to identify genes that are differentially expressed across conditions, highlighting those with large fold changes relative to their expression level.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ma_plot
#| code-fold: true
#| message: false

plotMA(contrast, alpha = 0.05, xlab = "Mean of normalized counts", ylab = "LFC")
```

# DE genes

DE genes are defined as those with an adjusted p-value below 0.05. Among them, upregulated genes are extracted as those with a LFC > 0.58, while downregulated genes are defined as those with an LFC < -0.58.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: de_genes
#| code-fold: true

up_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange > log2(1.5) & !is.na(padj)))
down_genes <- rownames(subset(contrast, padj < 0.05 & log2FoldChange < -log2(1.5) & !is.na(padj)))
```

# Volcano plot

A volcano plot displays each gene according to its LFC on the x-axis and the statistical significance (−log₁₀ adjusted p-value) on the y-axis. This time, there are a large number of DE chromosomal genes, both upregulated and downregulated, as well as some upregulated pBtic235 genes. Regarding GIL16, it has 22 upregulated genes among the 25 with the highest positive LFC, being the rest the chromosomal genes GBJ002_03410 (diacylglycerol kinase), GBJ002_01858 (glyoxalase/fosfomycin resistance/dioxygenase domain-containing protein) and GBJ002_02156 (bacterial transcription activator effector binding domain-containing protein). Furthermore, among the phage genes, there is no clear separation according to the promoter as occurred with GIL01, since all are observed to be differentially expressed and with LFC between 3.91 and 5.91. The only ones not differentially expressed are GIL16-DNT_01 (LFC = -0.76 and adjusted p-value = 0.7) and GIL16_13 (LFC = 5.58 and adjusted p-value = 0.135).

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: volcano_plot
#| code-fold: true

volcano_plot(contrast_df)
```

# Enrichment analysis

In this section, we present the results of enrichment analyses based on the COG (Clusters of Orthologous Groups), KEGG (Kyoto Encyclopedia of Genes and Genomes), GO (Gene Ontology) and EC (Enzyme Commission) categories for both upregulated and downregulated genes.

## COG

According to COG categories, we only found positive enrichment in energy production and conversion (**C**), replication, recombination and repair (**L**), chromatin structure and dynamics (**B**) and coenzyme transport and metabolism (**H**) due to 23, 18, 3 and 21 DE genes, respectively.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog
#| code-fold: true

cog_up <- cog_enrichment(up_genes, all_genes)
cog_down <- cog_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog_results
#| code-fold: true

datatable(cog_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(cog_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "COG", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

## KEGG

According to KEGG categories, we found positive and negative enrichments. In the case of the former, they were mainly related to metabolism, but also to replication and repair processes, specifically nucleotide excision repair, due to GBJ002_01735 (excinuclease ABC subunit C), GBJ002_02363 (excinuclease ABC subunit UvrA), GBJ002_02364 (excinuclease ABC subunit B), GBJ002_03098 (DNA helicase PcrA), GBJ002_03099 (NAD-dependent DNA ligase LigA) and GBJ002_03820 (RNA polymerase recycling motor HelD) genes. In the case of the latter, it resulted in depletion of pathways related to amino acids and membrane transport by ABC transporters.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg
#| message: false
#| code-fold: true

kegg_up <- kegg_enrichment(up_genes, all_genes)
kegg_down <- kegg_enrichment(down_genes, all_genes)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg_results
#| code-fold: true

datatable(kegg_up, rownames = FALSE, caption = "Positive enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(kegg_down, rownames = FALSE, caption = "Negative enrichment",
  extensions = "Scroller", options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 3) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## GO

The Gene Ontology (GO) is organized into three categories: **Biological Process** (BP), which describes the biological objectives to which gene products contribute; **Molecular Function** (MF), which refers to the elemental activities of gene products at the molecular level; and **Cellular Component** (CC), which specifies the locations within the cell where gene products are active.

-   According to **BP**, we observed positive enrichment in processes related to stress response and carbohydrates and nucleotides metabolism, while negative enrichment was associated to transport pathways and metabolic processes of dicarboxylic acids.

-   Regarding **MF**, we found positive enrichment in processes related to energy and transmembrane transporter activity of protons and inorganic and monoatomic ions, while that of oganic anions and organic and carboxylic acids was depleted. This could suggest a change in membrane potential.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go
#| code-fold: true
#| message: false

go_bp_up <- go_enrichment(up_genes, all_genes, "BP")
go_bp_down <- go_enrichment(down_genes, all_genes, "BP")

go_mf_up <- go_enrichment(up_genes, all_genes, "MF")
go_mf_down <- go_enrichment(down_genes, all_genes, "MF")

go_cc_up <- go_enrichment(up_genes, all_genes, "CC")
go_cc_down <- go_enrichment(down_genes, all_genes, "CC")
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go_results
#| code-fold: true

datatable(go_bp_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_bp_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_mf_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_up, rownames = FALSE, caption = "Positive enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

datatable(go_cc_down, rownames = FALSE, caption = "Negative enrichment", extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## EC

The EC classification establishes up to four levels to define enzymes. In this section, the first three were analyzed, achieving statistical significance in second and third levels for negative and positive enrichment, respectively. In the second level, we found among the cis-trans-isomerases (**5.2**) 3 out of 4 DE genes: GBJ002_00049 (foldase protein PrsA 4), GBJ002_03937 (peptidylprolyl isomerase PrsA) and GBJ002_04990 (foldase protein PrsA 3). In the third, **7.1.2** level was enriched by 7 out of 8 genes, corresponding to F0F1 ATP synthase subunits genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec
#| code-fold: true

ec_1_up <- ec_enrichment(up_genes, all_genes, 1)
ec_1_down <- ec_enrichment(down_genes, all_genes, 1)

ec_2_up <- ec_enrichment(up_genes, all_genes, 2)
ec_2_down <- ec_enrichment(down_genes, all_genes, 2)

ec_3_up <- ec_enrichment(up_genes, all_genes, 3)
ec_3_down <- ec_enrichment(down_genes, all_genes, 3)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec_results
#| code-fold: true

datatable(ec_1_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_1_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_2_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_up, rownames = FALSE, caption = "Positive enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

datatable(ec_3_down, rownames = FALSE, caption = "Negative enrichment", colnames = c(
  "EC number", "Class", "Subclass", "Sub-subclass", "DE", "No DE", "Genes", "Odds ratio", "p value", "Adjusted p-value"), extensions = "Scroller",
  options = list(
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

# Enrichment summary

This table contains a summary of the enrichments for each category. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: enrichment_summary
#| code-fold: true

es_up <- enrichment_summary(up_genes, categories = list(cog_up, kegg_up, go_bp_up, go_mf_up, go_cc_up, ec_1_up, ec_2_up, ec_3_up))
es_down <- enrichment_summary(down_genes, categories = list(cog_down, kegg_down, go_bp_down, go_mf_down, go_cc_down, ec_1_down, ec_2_down, ec_3_down))
es_de <- rbind(es_up, es_down)
rownames(es_de) <- c("UP", "DOWN")

datatable(es_de, colnames = c("DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 2),
            scrollX = TRUE,
            scroller = TRUE,
            deferRender = TRUE))
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

create_summary(contrast_name = "Time points 30 vs 0", rds_name = "gil16_t30_t0")
```

