---
title: GIL01 hierarchical clustering
---

```{r}
#| include: false

library("DT")
library("ggplot2")
library("heatmaply")
library("dplyr")
library("tidyr")
library("tibble")
library("forcats")
library("propr")
library("dendextend")
library("fpc")
library("clusterProfiler")
library("org.Bthuringiensis.eg.db")
```

```{r}
#| include: false

source("rnaseq_functions.R")
```

```{r}
#| include: false

counts <- readRDS(file = "gene_coexpression_counts.rds")
metadata <- readRDS(file = "coexpression_metadata.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")
```

# Obtain rho proportionality

Correlation shows how two variables vary together. In the case of RNAseq data, these correlations are inevitably relative, since we don't have the total number of reads, but rather a proportion of the total. Although algorithms like WGCNA employ Pearson correlation, this can lead to spurious associations, and more recent advances, especially in metagenomics, suggest the use of compositionally-aware methods. One such method is the rho (ρ) statistic, which can be used instead of correlation as the basis for familiar analyses and visualization methods, including coexpression networks and clustered heatmaps. In this case, we used rho proportionality to measure coexpression among GBJ002 (chromosome and pBtic235) and GIL01 genes and GIL01 samples at times 0, 10, 30 and 60, excluding 60-mock samples, since they are not induced by MMC and may introduce noise.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: rho
#| code-fold: true
#| message: false
#| warning: false

pseudocount <- min(counts[counts > 0])/10
counts <- counts + pseudocount
counts <- counts[1:4967, 13:24]
all_genes <- rownames(counts)

rhos <- propr(t(counts), metric = "rho")
```

# Dendrogram

The figure below shows a dendrogram representation of the hierarchical clustering generated by `hclust` with average method and at a distance (1 - ρ) of 0.25 (i.e., approximately 75% positive proportionality in expression). A higher cutoff height would yield larger clusters, although there might be genes with low coexpression within them. A lower cutoff height would yield greater proportionality in expression, although the clusters would be smaller. Furthermore, to simplify the analysis, only clusters with 20 or more genes were obtained.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: hc
#| code-fold: true
#| fig-width: 12
#| fig-height: 7

hc <- hclust(as.dist(1 - rhos@matrix), method = "average")

hc_clusters <- cutree(hc, h = 0.25)
cluster_sizes <- table(hc_clusters)
big_ids <- as.numeric(names(cluster_sizes[cluster_sizes >= 20]))
small_ids <- as.numeric(names(cluster_sizes[cluster_sizes < 20]))
leaf_order <- hc$order
clusters_in_order <- hc_clusters[leaf_order]
clusters_left_to_right <- rle(clusters_in_order)$values
which_big <- match(big_ids, clusters_left_to_right)
which_big <- which_big[!is.na(which_big)]

x_pos <- seq_along(leaf_order)
centers <- tapply(x_pos, clusters_in_order, mean)
centers_big <- centers[as.character(big_ids)]

genes_big <- names(hc_clusters)[hc_clusters %in% big_ids]
clusters_big <- hc_clusters[hc_clusters %in% big_ids]

cluster_gene_table <- data.frame(
  Cluster = clusters_big,
  Gene = genes_big,
  stringsAsFactors = FALSE
)

sizes <- cluster_gene_table %>% count(Cluster, name = "Size")
first_appearance <- cluster_gene_table %>% distinct(Cluster) %>% mutate(first_pos = row_number())

map_tbl <- sizes %>%
  left_join(first_appearance, by = "Cluster") %>%
  arrange(Size, first_pos) %>%
  group_by(Size) %>%
  mutate(
    tie_index = row_number(),
    letter = if (n() > 1) num_to_letters(tie_index) else rep("", n()),
    size_lab = sprintf("%03d", Size),
    new_name = paste0("GIL01_C", size_lab, letter)
  ) %>% ungroup()

cluster_to_name <- setNames(map_tbl$new_name, map_tbl$Cluster)

cluster_gene_table <- cluster_gene_table %>%
  mutate(Cluster = cluster_to_name[as.character(Cluster)]) %>%
  dplyr::left_join(
    annotation,
    by = c("Gene" = "Gene_ID")
  )

big_labels <- unname(cluster_to_name[as.character(big_ids)])

op <- par(col = "grey", xpd = NA, mar = par("mar") + c(1,0,0,0))
plot(hc, labels = FALSE,
     main = NA,
     sub = NA, xlab = NA, ylab = "Distance (1 - ρ)")
rect.hclust(hc, which = which_big, h = 0.25, border = which_big)

y_min <- par("usr")[3]
text(x = centers_big,
     y = y_min - 0.01 * diff(par("usr")[3:4]),
     labels = big_labels,
     cex = 0.5,
     srt = 90,
     adj = 1,
     col = "black")
```

```{r}
#| include: false

par(op)
```

# Stability

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: clusters_stability
#| code-fold: true

clusters_ids <- sort(c(big_ids, small_ids))

clusters_list <- lapply(clusters_ids, function(id) {
  names(hc_clusters)[hc_clusters == id]
})

names(clusters_list) <- clusters_ids

clusters_stability <- compute_stability(clusters = clusters_ids, strain = "GIL01", frac = 0.80, B = 1000)

datatable(clusters_stability, rownames = FALSE,
  filter = "top", extensions = c("Buttons", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatPercentage(columns = "Stability", digits = 2)
```

# Heatmap

After obtaining the dendrograms, we will represent how similar they are to each other with respect to gene expression patterns. To do this, we will use the distance of all genes in each cluster from the others to obtain an average distance (1 - ρ) and graphically represent the results in an interactive heatmap. As we can see, there is a main subdivision into two branches, among which we could distinguish some superclusters based on distance. For example, GIL01_C290 and GIL01_C026a have a distance (1 - ρ) of 0.2723, i.e., raising the threshold to this value would merge them into a supercluster. The next closest would be GIL01_C029b, which has distances of 0.2729 and 0.3357 to GIL01_C290 and GIL01_C026a, respectively. Depending on the enrichment results, we could consider raising the cutoff threshold to obtain separate functions in each cluster, although for the moment we will maintain the previously defined threshold.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: heatmap
#| code-fold: true
#| warning: false

labels_new <- unname(cluster_to_name[as.character(big_ids)])

members_list <- setNames(
  lapply(big_ids, function(k) which(hc_clusters == k)),
  labels_new
)

avg_inter <- matrix(0,
                    nrow = length(big_ids),
                    ncol = length(big_ids),
                    dimnames = list(labels_new, labels_new))

for (i in seq_len(length(big_ids))) {
  for (j in i:length(big_ids)) {
    if (i != j) {
      m <- mean((1 - rhos@matrix)[members_list[[i]], members_list[[j]]])
      avg_inter[i, j] <- m
      avg_inter[j, i] <- m
    }
  }
}

hc_row <- hclust(as.dist(avg_inter), method = "average")
hc_col <- hclust(as.dist(avg_inter), method = "average")

dend_row <- as.dendrogram(hc_row)
dend_col <- as.dendrogram(hc_col)

dend_row_olo <- seriate_dendrogram(dend_row, as.dist(avg_inter), method = "OLO")
dend_col_olo <- seriate_dendrogram(dend_col, as.dist(avg_inter), method = "OLO")

order_row <- order.dendrogram(dend_row_olo)
order_col <- order.dendrogram(dend_col_olo)

order_x <- colnames(avg_inter)[order_col]
order_y <- rownames(avg_inter)[order_row]

heatmaply(
  avg_inter[order_y, order_x, drop = FALSE],
  colors = viridis(n = 256, alpha = 1, begin = 1, end = 0, option = "viridis"),
  distfun_row = function(x) as.dist(x),
  distfun_col = function(x) as.dist(x),
  hclustfun_row = function(d) hclust(d, method = "average"),
  hclustfun_col = function(d) hclust(d, method = "average"),
  seriate = "none",
  symm = TRUE,
  scale = "none",
  label_names = c("Row", "Column", "Distance (1 - ρ)"),
  width = 1000,
  height = 800
)
```

# Clusters composition

The interactive table below displays clusters with 20 or more genes and the genes that comprise them. Above each column is a filter box that allows to filter by cluster, gene name, location, etc. At first glance, we observe that 2023 out of 5023 genes (40.27%) are located in clusters. Also, it can be observed that a large proportion of pBtic235 genes belong to cluster GIL01_C290. Regarding the prophage region (GBJ002_00964 to GBJ002_01009 genes), many of the genes are in GIL01_C055. Last but not least, GIL01 genes are mostly found in GIL01_C055.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cluster_gene_table
#| code-fold: true

cluster_gene_table$Cluster <- factor(cluster_gene_table$Cluster, levels = order_x)
cluster_gene_table$Element <- factor(cluster_gene_table$Element, levels = c("chromosome", "pBtic235", "GIL01"))
cluster_gene_table$Strand <- factor(cluster_gene_table$Strand, levels = c("+", "-"))

datatable(cluster_gene_table, rownames = FALSE,
  colnames = c("Cluster", "Gene", "Element", "Start", "End", "Strand", "Gene symbol", "Description", "COG category", "Association", "Type", "Subtype", "Source"),
  filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 2),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE))
```

# Expression plots

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: expression_plots
#| code-fold: true
#| warning: false

expression_plots <- clusters_expression(expression_matrix = rhos@logratio,
                                        metadata = metadata, times = c(0, 10, 30, 60),
                                        de_genes = c("gil01_t10_t0", "gil01_t30_t0", "gil01_t60_t0", "gil01_t60_t60-mock", "gil01_t60-mock_t0"),
                                        which_contrast = 3)
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: all_clusters
#| code-fold: true
#| fig-width: 14
#| fig-height: 8

expression_plots$all_clusters
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: facet_clusters
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

expression_plots$facet_clusters
```

# Enrichment analysis

As with determining the enrichment of upregulated or downregulated genes, an enrichment analysis was performed according to the COG, KEGG, GO and EC categories to determine if clusters are enriched in any function.

## COG

Regarding COG categories, 9 clusters show significant enrichment comprising 7 distinct terms. Among the most notable findings are GIL01_C032b, which is enriched in genes related to cell motility (**N**), and GIL01_C029a, enriched in mobilome: prophages, transposons (**X**) category.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: cog
#| code-fold: true

cog_results <- cog_enrichment_by_cluster(cluster_gene_table, all_genes)

datatable(cog_results, rownames = FALSE,
  colnames = c("COG", "IN", "OUT", "Genes", "Odds ratio", "p value", "Adjusted p-value", "Cluster"),
  extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 0, rightColumns = 1),
    scrollX = TRUE,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

## KEGG

Regarding KEGG terms, there is a wide variety across the clusters. GIL01_C032b is again related to cell motility pathways, while GIL01_C029b, GIL01_C020c and GIL01_C020d are realted to membrane transport pathways. Regarding replication and repair pathways, we can observe two distinct clusters: GIL01_C290, associated to DNA replication, homologous recombination and mismatch repair, and GIL01_C036c, associated to nucleotide excision repair. Interestingly, 9 clusters appears to be enriched in diverse metabolic pathways, which suggests a specialization in gene expression in the different reflected pathways.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: kegg
#| code-fold: true
#| warning: false

kegg_results <- kegg_enrichment_by_cluster(cluster_gene_table, all_genes)

datatable(kegg_results, rownames = FALSE,
  colnames = c("Category", "Subcategory", "Pathway ID", "Pathway description", "Gene ratio",
  "Background ratio", "Rich factor", "Fold enrichment", "Z score", "p value", "Adjusted p‑value",
  "q value", "Genes", "Gene count", "Cluster"),
  extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 0, rightColumns = 1),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## GO

GO enrichment analysis is much more complex, since the presence of few genes can result in different enrichment levels. In any case, three independent tables (BP, MF and CC) have been generated and are summarized in the Enrichment summary section.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: go
#| code-fold: true
#| warning: false

go_bp_results <- go_enrichment_by_cluster(cluster_gene_table, all_genes, "BP")
datatable(go_bp_results, rownames = FALSE,
  colnames = c("GO ID", "GO term description", "Gene ratio", "Background ratio",
  "Rich factor", "Fold enrichment", "Z score", "p value", "Adjusted p‑value",
  "q value", "Genes", "Gene count", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "BP enrichment"),
  extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 0, rightColumns = 1),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

go_mf_results <- go_enrichment_by_cluster(cluster_gene_table, all_genes, "MF")
datatable(go_mf_results, rownames = FALSE,
  colnames = c("GO ID", "GO term description", "Gene ratio", "Background ratio",
  "Rich factor", "Fold enrichment", "Z score", "p value", "Adjusted p‑value",
  "q value", "Genes", "Gene count", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "MF enrichment"),
  extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 0, rightColumns = 1),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)

go_cc_results <- go_enrichment_by_cluster(cluster_gene_table, all_genes, "CC")
datatable(go_cc_results, rownames = FALSE,
  colnames = c("GO ID", "GO term description", "Gene ratio", "Background ratio",
  "Rich factor", "Fold enrichment", "Z score", "p value", "Adjusted p‑value",
  "q value", "Genes", "Gene count", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "CC enrichment"),
  extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 0, rightColumns = 1),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE)) %>%
  formatRound(columns = c("RichFactor", "FoldEnrichment", "zScore"), digits = 2) %>%
  formatSignif(columns = c("pvalue", "p.adjust", "qvalue"), digits = 3)
```

## EC

Although there are not many EC terms, some interesting results can be observed. For example, GIL01_C290 is enriched at each of numbers **1**, **1.6** and **1.6.5**, which correspond to NADH-quinone oxidoreductase subunits genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: ec
#| code-fold: true
#| warning: false

ec_1_results <- ec_enrichment_by_cluster(cluster_gene_table, all_genes, 1)
datatable(ec_1_results, rownames = FALSE,
  colnames = c("EC number", "Class", "IN", "OUT", "Genes", "Odds ratio",
  "p value", "Adjusted p-value", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "EC level 1 enrichment"),
  extensions = "Buttons", options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"))) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

ec_2_results <- ec_enrichment_by_cluster(cluster_gene_table, all_genes, 2)
datatable(ec_2_results, rownames = FALSE,
  colnames = c("EC number", "Class", "Subclass", "IN", "OUT", "Genes", "Odds ratio",
  "p value", "Adjusted p-value", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "EC level 2 enrichment"),
  extensions = "Buttons", options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"))) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)

ec_3_results <- ec_enrichment_by_cluster(cluster_gene_table, all_genes, 3)
datatable(ec_3_results, rownames = FALSE,
  colnames = c("EC number", "Class", "Subclass", "Sub-subclass", "IN", "OUT", "Genes", "Odds ratio",
  "p value", "Adjusted p-value", "Cluster"),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: start;",
    "EC level 3 enrichment"),
  extensions = "Buttons", options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"))) %>%
  formatRound(columns = "odds.ratio", digits = 2) %>%
  formatSignif(columns = c("p.value", "padj"), digits = 3)
```

# Enrichment summary

This table contains a summary of the enrichments for each category. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: enrichment_summary
#| code-fold: true

size_df <- as.data.frame(table(cluster_gene_table$Cluster)) %>%
  dplyr::rename(Cluster = Var1, Size = Freq)

cog_df <- cog_results %>% group_by(Cluster) %>%
  dplyr::summarise(COG = collapse_vals(COG), .groups = "drop")

kegg_df <- kegg_results %>% group_by(Cluster) %>%
  dplyr::summarise(KEGG = collapse_vals(subcategory), .groups = "drop")

go_bp_df <- go_bp_results %>% group_by(Cluster) %>%
  dplyr::summarise(`GO-BP`= collapse_vals(Description), .groups = "drop")

go_mf_df <- go_mf_results %>% group_by(Cluster) %>%
  dplyr::summarise(`GO-MF`= collapse_vals(Description), .groups = "drop")

go_cc_df <- go_cc_results %>% group_by(Cluster) %>%
  dplyr::summarise(`GO-CC`= collapse_vals(Description), .groups = "drop")

ec_1_df <- ec_1_results %>% group_by(Cluster) %>%
  dplyr::summarise(`EC-1` = collapse_vals(EC), .groups = "drop")

ec_2_df <- ec_2_results %>% group_by(Cluster) %>%
  dplyr::summarise(`EC-2` = collapse_vals(EC), .groups = "drop")

ec_3_df <- ec_3_results %>% group_by(Cluster) %>%
  dplyr::summarise(`EC-3` = collapse_vals(EC), .groups = "drop")

summary_df <- size_df %>%
  dplyr::left_join(cog_df,   by = "Cluster") %>%
  dplyr::left_join(kegg_df,  by = "Cluster") %>%
  dplyr::left_join(go_bp_df, by = "Cluster") %>%
  dplyr::left_join(go_mf_df, by = "Cluster") %>%
  dplyr::left_join(go_cc_df, by = "Cluster") %>%
  dplyr::left_join(ec_1_df,   by = "Cluster") %>%
  dplyr::left_join(ec_2_df,   by = "Cluster") %>%
  dplyr::left_join(ec_3_df,   by = "Cluster") %>%
  dplyr::select(Cluster, Size, COG, KEGG, `GO-BP`, `GO-MF`, `GO-CC`, `EC-1`, `EC-2`, `EC-3`) %>%
  dplyr::mutate(Cluster = factor(Cluster, levels = levels(cluster_gene_table$Cluster)))

datatable(summary_df, rownames = FALSE,
  filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
    autoWidth = TRUE,
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 2),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE))
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: results
#| code-fold: true

gil01_clusters <- split(cluster_gene_table$Gene, cluster_gene_table$Cluster)

saveRDS(gil01_clusters, file = "gil01_clusters.rds")
saveRDS(clusters_stability, file = "gil01_clusters_stability.rds")
```
