---
title: GIL16 lytic cycle induction analysis
---

```{r}
#| include: false

library("DT")
library("edgeR")
library("DESeq2")
library("ggplot2")
library("plotly")
library("heatmaply")
library("pheatmap")
library("dplyr")
library("tidyr")
library("clusterProfiler")
library("RColorBrewer")
library("org.Bthuringiensis.eg.db")
```

```{r}
#| include: false

source("rnaseq_functions.R")
```

# Load data

In this section, we load the data necessary to begin the RNAseq analysis: counts and metadata. We also load genome annotation files to visualize the results and the rRNA genes, which are one of the most common sources of noise in RNAseq analyses. Although rRNA gene reads removal was efficient using Bowtie2, we chose to discard these gene counts before normalization.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: load_data
#| code-fold: true
#| warning: false

counts <- readRDS(file = "lytic_cycle_induction_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_metadata.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")
```

# Delete GIL01 genes

Since there are no GIL01 genes in the GIL16 samples, we removed them before creating the dds object.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: delete_genes
#| code-fold: true
#| message: false
#| warning: false

counts <- counts[!grepl("^GIL01", rownames(counts)), , drop = FALSE]
counts <- counts[, grepl("^GIL16", colnames(counts)), drop = FALSE]

all_genes <- rownames(counts)
```

# Create dds object

With the following code, we modified the metadata file and created the dds object using a design that took into account both time and replication to model gene expression.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dds
#| code-fold: true
#| message: false
#| warning: false

metadata <- metadata[grepl("^GIL16", metadata$Strain), ]
metadata$Strain <- NULL

counts[] <- lapply(counts, function(x) if(is.numeric(x)) ifelse(x %% 1 == 0.5, ceiling(x), round(x)) else x)

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time)

dds <- DESeq(dds)
```

# Dimensionality reduction

DESeq2 uses Variance Stabilizing Transformation (VST) to transform RNAseq counts so that the variance is approximately constant across different expression levels. This facilitates analyses like clustering or dimensionality reduction techniques, such as PCA and MDS.

The results obtained are good and highly consistent between PCA and MDS. The first two dimensions explain more than 55% of the variability in gene expression (or separation between samples). As observed, 0 and 10 samples tend to cluster together, while 30 and 60 samples are better separated along the first dimension. In contrast, 60-mock samples differ from the others along the second dimension. 

When examining the distribution of samples, replicate I tends to occupy lower positions along the second dimension. This pattern suggests the presence of the batch effect described earlier. However, previous analyses indicate that this effect is not a major concern as long as expression is modeled with Replicate included as a covariate. Additionally, it is worth noting that replicate I at time 60 lies noticeably farther from the other two replicates along the second dimension, suggesting that the batch effect may become more pronounced over time.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dimensionality_reduction
#| message: false
#| warning: false
#| code-fold: true
#| fig-width: 7
#| fig-height: 5

vsdata <- vst(dds, blind = TRUE)
pca_df <- plotPCA(vsdata, intgroup = "Time", ntop = nrow(dds), returnData = TRUE)
pca_df$Sample <- rownames(pca_df)

percentVar <- round(100 * attr(pca_df, "percentVar"), 2)

pca_df$hover <- paste0(
  "Sample: ", pca_df$Sample, "<br>",
  "PC1: ", round(pca_df$PC1, 2), "<br>",
  "PC2: ", round(pca_df$PC2, 2)
)

p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Time, text = hover)) +
  geom_point(size = 3) +
  labs(
    x = paste0("PC1 (", percentVar[1], "%)"),
    y = paste0("PC2 (", percentVar[2], "%)"),    
    color = "Time") +
  theme_minimal()

ggplotly(p, tooltip = "text")

dge <- DGEList(counts = counts(dds))
dge <- calcNormFactors(dge)
mds <- plotMDS(dge, top = nrow(dds), plot = FALSE)

mds_df <- data.frame(
  MDS1 = mds$x,
  MDS2 = mds$y,
  Sample = colnames(dge),
  Time = colData(dds)$Time
)

percentVar <- round(100 * mds$var.explained, 2)

mds_df$hover <- paste0(
  "Sample: ", mds_df$Sample, "<br>",
  "MDS1: ", round(mds_df$MDS1, 2), "<br>",
  "MDS2: ", round(mds_df$MDS2, 2)
)

p <- ggplot(mds_df, aes(x = MDS1, y = MDS2, color = Time, text = hover)) +
  geom_point(size = 3) +
  labs(
    x = paste0("MDS1 (", percentVar[1], "%)"),
    y = paste0("MDS2 (", percentVar[2], "%)"),
    color = "Group"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

Below this plot, we show two heatmaps with the 20 genes that contribute most to PC1 and PC2 in the PCA.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: genes_pca
#| code-fold: true

vs_mat <- assay(vsdata)
topVarGenes <- head(order(rowVars(vs_mat), decreasing = TRUE), nrow(dds))
pca <- prcomp(t(vs_mat[topVarGenes, ]))
sample_pca <- as.data.frame(pca$x)
sample_pca$Time <- colData(vsdata)$Time

gene_loadings <- as.data.frame(pca$rotation)
top_genes_PC1 <- head(gene_loadings[order(abs(gene_loadings$PC1), decreasing=TRUE), ], 20)
top_genes_PC2 <- head(gene_loadings[order(abs(gene_loadings$PC2), decreasing=TRUE), ], 20)

labels_PC1 <- mark_pBtic235_genes(rownames(top_genes_PC1), annotation)
labels_PC2 <- mark_pBtic235_genes(rownames(top_genes_PC2), annotation)
```

## PC1

In the case of PC1, we can observe a clear separation into three parts: samples induced at 60 minutes, samples induced at early times and without induction and samples induced at 30 minutes. GIL16_t10_III is grouped with the rest of the 30 minutes samples, as seen in the PCA plot. Regarding the 20 most variable genes, as expected, they correspond to GIL16 genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: pheatmap_pc1
#| code-fold: true
#| fig-width: 10
#| fig-height: 7

pheatmap(vs_mat[rownames(top_genes_PC1), ], labels_row = labels_PC1, legend = FALSE)
```

## PC2

In the case of PC2, we found less separation between the times. Replicates II and III at times 0 and 10 are grouped in one branch, but replicates I at those times are more closely associated with replicates at time 30. A similar pattern is observed with times 30 and 60, as replicate I at time 30 is associated with 60-mock samples and GIL16_t60_I. In any case, the grouping of samples induced at 60 minutes with others at 10 and 30 minutes seems strange. Although it may seem contradictory, PC1 explains much more variance in gene expression than PC2 and 20 genes might prove insufficient to properly cluster samples in this component. To clarify these differences, we will examine the clustering between samples in the heatmap.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: pheatmap_pc2
#| code-fold: true
#| fig-width: 10
#| fig-height: 7

pheatmap(vs_mat[rownames(top_genes_PC2), ], labels_row = labels_PC2, legend = FALSE)
```

# Heatmap

The heatmap is divided into two main branches: samples induced at 60 minutes and the rest. The former are much more distinct, even from each other, than from the remaining induced and non-induced samples. Among the remaining samples, the non-induced and induced samples (induced at 30 minutes) cluster appropriately.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: heatmap
#| code-fold: true
#| warning: false
#| fig-width: 10
#| fig-height: 7

sampleDists <- dist(t(assay(vsdata)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsdata$Sample)

heatmaply(
  sampleDistMatrix,
  colors = viridis(n = 256, alpha = 1, begin = 1, end = 0, option = "viridis"),
  Rowv = as.dendrogram(hclust(sampleDists)),
  Colv = as.dendrogram(hclust(sampleDists)),
  label_names = c("Row", "Column", "Value")
)
```

# Boxplot and dispersion plot

The boxplot of VST-transformed expression values illustrates the distribution of normalized expression levels across samples. All samples show very similar medians, interquartile ranges and overall dispersion, with no evidence of systematic shifts between experimental groups. This indicates that the variance-stabilizing transformation has effectively reduced heteroscedasticity and that the normalization procedure has removed global differences in expression distributions, supporting the suitability of the data for downstream analyses. If anything had to be pointed out, it seems that the induced samples from minute 60 show a greater number of outliers.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: boxplot
#| code-fold: true
#| message: false
#| fig-width: 10
#| fig-height: 7

op <- par(mar = c(10, 4, 4, 2))
boxplot(assay(vsdata), las = 2)
```

```{r}
#| include: false

par(op)
```

The dispersion versus mean of normalized counts plot illustrates the mean–variance relationship assumed by the negative binomial model. Gene-wise dispersion estimates (black points) show higher variability at low expression levels and progressively lower dispersion as mean expression increases. The fitted dispersion trend (red line) captures this overall pattern well and the final dispersion estimates after shrinkage (blue points) are stabilized around this curve. The close agreement between observed and fitted dispersions, together with the lack of anomalous structure, suggests that the dispersion estimation is robust and that the model assumptions are appropriate for these data.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dispersion
#| code-fold: true
#| message: false
#| fig-width: 10
#| fig-height: 7

plotDispEsts(dds)
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

saveRDS(counts, file = "lytic_cycle_induction_gil16_counts.rds")
saveRDS(metadata, file = "lytic_cycle_induction_gil16_metadata.rds")
saveRDS(all_genes, file = "lytic_cycle_induction_gil16_all_genes.rds")
```
