---
title: GBJ002 lytic cycle induction analysis
---

```{r}
#| include: false

library("DT")
library("edgeR")
library("DESeq2")
library("ggplot2")
library("plotly")
library("heatmaply")
library("pheatmap")
library("dplyr")
library("tidyr")
library("clusterProfiler")
library("RColorBrewer")
library("org.Bthuringiensis.eg.db")
```

```{r}
#| include: false

source("tectivirnaseq_functions.R")
```

# Load data

In this section, we load the data necessary to begin the RNAseq analysis: counts and metadata. We also load genome annotation files to visualize the results.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: load_data
#| code-fold: true
#| warning: false

counts <- readRDS(file = "lytic_cycle_induction_counts.rds")
metadata <- readRDS(file = "lytic_cycle_induction_metadata.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")
```

# Delete phages genes

Since there are no phage genes in the GBJ002 samples, we removed them before creating the dds object.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: delete_genes
#| code-fold: true
#| message: false
#| warning: false

counts <- counts[!grepl("^GIL", rownames(counts)), , drop = FALSE]
counts <- counts[, !grepl("^GIL", colnames(counts)), drop = FALSE]

all_genes <- rownames(counts)
```

# Create dds object

With the following code, we modified the metadata file and created the dds object using a design that took into account both time and replication to model gene expression.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dds
#| code-fold: true
#| message: false
#| warning: false

metadata <- metadata[!grepl("^GIL", metadata$Strain), ]
metadata$Strain <- NULL

counts[] <- lapply(counts, function(x) if(is.numeric(x)) ifelse(x %% 1 == 0.5, ceiling(x), round(x)) else x)

dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~ Replicate + Time + MMC)

dds <- DESeq(dds)
```

# Dimensionality reduction

DESeq2 uses Variance Stabilizing Transformation (VST) to transform RNAseq counts so that the variance is approximately constant across different expression levels. This facilitates analyses like clustering or dimensionality reduction techniques, such as PCA and MDS.

The results obtained are good and highly consistent between PCA and MDS. The first two dimensions explain almost 70% of the variability in gene expression (or separation between samples). As observed, 0 and 10 samples tend to cluster together, while 30 and 60 samples are better separated along the first dimension. In contrast, 60-mock samples differ from the others along the second dimension. 

When examining the distribution of samples, replicates I, III and II appear to occupy decreasing positions along the second dimension for the 0, 10, 30 and 60 samples, whereas this pattern is less clear for the 60-mock samples. This suggests the presence of the batch effect described earlier. Nevertheless, previous results indicate that this effect is not a major concern provided that expression is modeled with Replicate included as a covariate.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dimensionality_reduction
#| message: false
#| warning: false
#| code-fold: true
#| fig-width: 7
#| fig-height: 5

vsdata <- vst(dds, blind = TRUE)
pca_df <- plotPCA(vsdata, intgroup = "Time", ntop = nrow(dds), returnData = TRUE)
pca_df$Sample <- rownames(pca_df)

percentVar <- round(100 * attr(pca_df, "percentVar"), 2)

pca_df$hover <- paste0(
  "Sample: ", pca_df$Sample, "<br>",
  "PC1: ", round(pca_df$PC1, 2), "<br>",
  "PC2: ", round(pca_df$PC2, 2)
)

p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Time, text = hover)) +
  geom_point(size = 3) +
  labs(
    x = paste0("PC1 (", percentVar[1], "%)"),
    y = paste0("PC2 (", percentVar[2], "%)"),    
    color = "Time") +
  theme_minimal()

ggplotly(p, tooltip = "text")

dge <- DGEList(counts = counts(dds))
dge <- calcNormFactors(dge)
mds <- plotMDS(dge, top = nrow(dds), plot = FALSE)

mds_df <- data.frame(
  MDS1 = mds$x,
  MDS2 = mds$y,
  Sample = colnames(dge),
  Time = colData(dds)$Time
)

percentVar <- round(100 * mds$var.explained, 2)

mds_df$hover <- paste0(
  "Sample: ", mds_df$Sample, "<br>",
  "MDS1: ", round(mds_df$MDS1, 2), "<br>",
  "MDS2: ", round(mds_df$MDS2, 2)
)

p <- ggplot(mds_df, aes(x = MDS1, y = MDS2, color = Time, text = hover)) +
  geom_point(size = 3) +
  labs(
    x = paste0("MDS1 (", percentVar[1], "%)"),
    y = paste0("MDS2 (", percentVar[2], "%)"),
    color = "Group"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

Below this plot, we show two heatmaps with the 20 genes that contribute most to PC1 and PC2 in the PCA.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: genes_pca
#| code-fold: true

vs_mat <- assay(vsdata)
topVarGenes <- head(order(rowVars(vs_mat), decreasing = TRUE), nrow(dds))
pca <- prcomp(t(vs_mat[topVarGenes, ]))
sample_pca <- as.data.frame(pca$x)

gene_loadings <- as.data.frame(pca$rotation)
top_genes_PC1 <- head(gene_loadings[order(abs(gene_loadings$PC1), decreasing=TRUE), ], 20)
top_genes_PC2 <- head(gene_loadings[order(abs(gene_loadings$PC2), decreasing=TRUE), ], 20)

labels_PC1 <- mark_pBtic235_genes(rownames(top_genes_PC1), annotation)
labels_PC2 <- mark_pBtic235_genes(rownames(top_genes_PC2), annotation)
```

## PC1

In the case of PC1, we see that all samples cluster according to their time, except for replicate III at minute 10, which is more similar to those at time 0. In any case, the most different samples are those at time 60. Regarding the genes, we see a clear difference in the genes of the prophage region: expression is much higher at minute 60 induced samples than in the other groups.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: pheatmap_pc1
#| code-fold: true
#| fig-width: 10
#| fig-height: 7

pheatmap(vs_mat[rownames(top_genes_PC1), ], labels_row = labels_PC1, legend = FALSE)
```

## PC2

In the case of PC2, all samples cluster according to their time, but this time there is a separation between the 60 and 60-mock samples from the rest, as seen in PCA. The greatest difference in gene expression is found in the genes GBJ002_01175 to GBJ002_01179, which encode two phosphoesterases, two SR1 sRNA and one SR1 protein.

According to [RFAM](https://rfam.org/family/RF02376), SR1 RNA is a small RNA (sRNA) produced by species of *Bacillus* and closely related bacteria whose expression is regulated by CcpA and CcpN. It is a dual-function RNA which acts both as a protein-coding RNA and as a regulatory sRNA:

-   SR1 RNA is involved in the regulation of arginine catabolism. SR1 RNA binds to complementary stretches of ahrC mRNA (also known as argR) and inhibits translation. AhrC endodes an arginine repressor protein which represses synthesis of arginine biosynthetic enzymes and activates arginine catabolic enzymes via regulation of the rocABC and rocDEF operons.

-   In addition to acting as a sRNA, SR1 also encodes a small peptide, SR1P. SR1P binds to glyceraldehyde-3-phosphate dehydrogenase (GapA) and stabilises the gapA operon mRNAs.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: pheatmap_pc2
#| code-fold: true
#| fig-width: 10
#| fig-height: 7

pheatmap(vs_mat[rownames(top_genes_PC2), ], labels_row = labels_PC2, legend = FALSE)
```

# Heatmap

The heatmap uses information from all genes in the model built using VST transformation to see how the samples cluster. A separation into three zones is observed: samples of times 0 and 10 minutes, 60-mock samples and samples of times 30 and 60. In the first zone it is observed that the times do not separate the samples well, since replicates I and II are being associated. In any case, this is not a cause for concern.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: heatmap
#| code-fold: true
#| warning: false
#| fig-width: 10
#| fig-height: 7

sampleDists <- dist(t(assay(vsdata)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsdata$Sample)

heatmaply(
  sampleDistMatrix,
  colors = viridis(n = 256, alpha = 1, begin = 1, end = 0, option = "viridis"),
  Rowv = as.dendrogram(hclust(sampleDists)),
  Colv = as.dendrogram(hclust(sampleDists)),
  label_names = c("Row", "Column", "Value")
)
```

# Boxplot and dispersion plot

The boxplot of VST-transformed expression values illustrates the distribution of normalized expression levels across samples. All samples show very similar medians, interquartile ranges and overall dispersion, with no evidence of systematic shifts between experimental groups. This indicates that the variance-stabilizing transformation has effectively reduced heteroscedasticity and that the normalization procedure has removed global differences in expression distributions, supporting the suitability of the data for downstream analyses.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: boxplot
#| code-fold: true
#| message: false
#| fig-width: 10
#| fig-height: 7

op <- par(mar = c(10, 4, 4, 2))
boxplot(assay(vsdata), las = 2)
```

```{r}
#| include: false

par(op)
```

The dispersion versus mean of normalized counts plot illustrates the mean–variance relationship assumed by the negative binomial model. Gene-wise dispersion estimates (black points) show higher variability at low expression levels and progressively lower dispersion as mean expression increases. The fitted dispersion trend (red line) captures this overall pattern well and the final dispersion estimates after shrinkage (blue points) are stabilized around this curve. The close agreement between observed and fitted dispersions, together with the lack of anomalous structure, suggests that the dispersion estimation is robust and that the model assumptions are appropriate for these data.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: dispersion
#| code-fold: true
#| message: false
#| fig-width: 10
#| fig-height: 7

plotDispEsts(dds)
```

# Save results

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: save_results
#| code-fold: true

saveRDS(counts, file = "lytic_cycle_induction_gbj002_counts.rds")
saveRDS(metadata, file = "lytic_cycle_induction_gbj002_metadata.rds")
```