---
title: Results summary
---

```{r}
#| include: false

library("dplyr")
library("tidyr")
library("ggplot2")
library("plotly")
library("DT")
library("stringr")
library("patchwork")
library("ggVennDiagram")
library("htmltools")
library("UpSetR")
```

```{r}
#| include: false

source("tectivirnaseq_functions.R")
```

This page compiles the results of all the performed contrasts: in lysogeny, intra-strain time-course and inter-strain time-course.

# Load data

In this code chunk, we load the RDS objects, the counts and the annotation file.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: load_summaries
#| code-fold: true

load_summaries()
build_functional_enrichments()

lysogeny_counts <- readRDS(file = "lysogeny_counts.rds")
lytic_cycle_induction_counts <- readRDS(file = "lytic_cycle_induction_counts.rds")
annotation <- readRDS(file = "GBJ002-GIL01-GIL16.rds")
```

# Lysogeny summaries

This section presents the results of the lysogeny contrasts. The two bar charts show the total number of DE genes and the number of enrichments in each contrast. In addition, the enrichments in the COG and KEGG categories are represented in two bubble plots to facilitate comparison between contrasts. Finally, the interactive table compiles all the results from the **Enrichment summary** tables for each contrast.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_summaries
#| code-fold: true

lysogeny_contrasts <- combine_summaries(type = "contrast", group = "strains", filter = "Lysogeny")
lysogeny_enrichments <- combine_summaries(type = "enrichment", group = "strains", filter = "Lysogeny")
```

## Barplots

This plot shows the number of DE genes, both upregulated and downregulated, in each contrast. As we can see, the GIL01 vs GBJ002 contrast has the fewest DE genes, while the GIL16 vs GBJ002 contrast has the most.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_cs
#| code-fold: true
#| warning: false
#| fig-width: 12
#| fig-height: 6

lysogeny_contrasts_size <- lysogeny_contrasts %>%
  select(Contrast, DE, Size) %>%
  mutate(
    Contrast = factor(Contrast, levels = levels(lysogeny_contrasts$Contrast)),
    DE = factor(DE, levels = levels(lysogeny_contrasts$DE))
  )

ggplot(lysogeny_contrasts_size, aes(x = Contrast, y = Size, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = Size),
    position = position_dodge(width = 0.8),
    vjust = -0.4,
    size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  ylim(0, 600) +
  theme_minimal(base_size = 14) +
  labs(x = NULL, y = "DE genes", fill = "DE")
```

This plot shows the number of enrichments found in each category. Most enrichments are found in the MF and CC categories of GO, while the fewest are found in the EC numbers. Again, the GIL01 vs GBJ002 contrast has the fewest enrichments, while the GIL16 vs GBJ002 contrast has the most.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_ce
#| code-fold: true
#| warning: false
#| fig-width: 10
#| fig-height: 6

lysogeny_contrasts_enrichments <- lysogeny_contrasts %>%
  pivot_longer(
    cols = !c(Group, Contrast, DE),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  filter(Category != "Size") %>%
  mutate(
    Contrast = factor(Contrast, levels = levels(lysogeny_contrasts$Contrast)),
    DE = factor(DE, levels = levels(lysogeny_contrasts$DE)),
    Category = factor(Category, levels = colnames(lysogeny_contrasts)[4:ncol(lysogeny_contrasts)])
  )

ggplot(lysogeny_contrasts_enrichments, aes(x = Category, y = Count, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = ifelse(Count == 0, NA, Count)),
    position = position_dodge(width = 0.8),
    vjust = -0.4, size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  scale_x_discrete(labels = c("COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)")) +
  facet_wrap(~ Contrast, nrow = 3) +
  scale_y_continuous(limits = c(0, 40), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold")) +
  labs(x = NULL, y = "Enrichments", fill = "DE")
```

## Venn diagrams

Venn diagrams were used to evaluate the overlap of differentially expressed genes associated with lysogeny among the three pairwise strain comparisons. Separate analyses were performed for upregulated and downregulated genes. Each circle represents the set of genes showing differential expression in a given strain comparison, whereas overlapping regions indicate genes shared between two or more comparisons. Genes located in the intersections correspond to transcriptional differences that are common across strain contrasts, while genes found exclusively in a single region represent comparison-specific responses. This analysis facilitates the identification of both conserved and unique transcriptional signatures associated with lysogeny and provides an overview of the extent to which differential gene expression patterns are shared among strains.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: venn_lysogeny
#| code-fold: true
#| warning: false
#| fig-width: 8
#| fig-height: 5

venn_contrasts(group = "Lysogeny", de_genes = "UP", label_alpha = 0)
venn_contrasts(group = "Lysogeny", de_genes = "DOWN", label_alpha = 0.25)
```

## Bubble plots

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_cog
#| code-fold: true
#| fig-width: 10
#| fig-height: 6

cog_names <- read.csv("cog_name.tsv", sep = "\t", header = TRUE)

lysogeny_cog <- lysogeny_cog %>%
  rename(Number = DE) %>%
  mutate(
    Group = Contrast,
    DE = Direction,
    COG_letter = str_extract(COG, "\\(([A-Z])\\)") %>% str_remove_all("[\\(\\)]"),
    minus_log10_p = -log10(p)
  )

category_order <- cog_names %>%
  filter(COG %in% lysogeny_cog$COG_letter) %>%
  pull(COG)

ordered_categories <- lysogeny_cog %>%
  mutate(COG_letter = factor(COG_letter, levels = category_order)) %>%
  arrange(desc(COG_letter)) %>%
  distinct(COG) %>%
  pull(COG)

lysogeny_cog$COG <- factor(lysogeny_cog$COG, levels = ordered_categories)
lysogeny_cog$Group <- factor(lysogeny_cog$Group, levels = c("Strains GIL01 vs GBJ002", "Strains GIL16 vs GBJ002", "Strains GIL16 vs GIL01"))

lysogeny_cog <- lysogeny_cog %>%
  mutate(
    y_pos = as.numeric(COG) + ifelse(DE == "UP", 0.1, -0.1)
  )

line_data <- lysogeny_cog %>%
  distinct(COG, DE)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = as.numeric(COG) + 0.1),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = as.numeric(COG) - 0.1),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 0.35, y = as.numeric(COG) + 0.1, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 0.35, y = as.numeric(COG) - 0.1, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = lysogeny_cog,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      shape = Group
    )
  ) +
  scale_x_continuous(
    limits = c(0, 0.35),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_y_continuous(
    breaks = seq_along(levels(lysogeny_cog$COG)),
    labels = levels(lysogeny_cog$COG)
  ) +
  scale_fill_viridis_c(limits = c(0, 20)) +
  scale_size_continuous(
    limits = c(1, 100),
    breaks = c(0, 25, 50, 75, 100)
  ) +
  scale_shape_manual(
    values = c(
      "Strains GIL01 vs GBJ002" = 21,
      "Strains GIL16 vs GBJ002" = 22,
      "Strains GIL16 vs GIL01" = 23
    )
  ) +
  labs(
    x = "Gene ratio",
    y = "COG category",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_kegg
#| code-fold: true
#| fig-width: 10
#| fig-height: 6

btn_pathways <- read.csv("btn_pathways.tsv", sep = "\t", header = TRUE)

lysogeny_kegg <- lysogeny_kegg %>%
  rename(Number = DE) %>%
  mutate(
    Group = Contrast,
    DE = Direction,
    minus_log10_p = -log10(p)
  )

category_order <- btn_pathways %>%
  pull(KEGG.pathway.maps)

ordered_categories <- lysogeny_kegg %>%
  mutate(KEGG = factor(KEGG, levels = category_order)) %>%
  arrange(desc(KEGG)) %>%
  distinct(KEGG) %>%
  pull(KEGG)

lysogeny_kegg$KEGG <- factor(lysogeny_kegg$KEGG, levels = ordered_categories)
lysogeny_kegg$Group <- factor(lysogeny_kegg$Group, levels = c("Strains GIL01 vs GBJ002", "Strains GIL16 vs GBJ002", "Strains GIL16 vs GIL01"))

lysogeny_kegg <- lysogeny_kegg %>%
  mutate(
    y_pos = as.numeric(KEGG) + ifelse(DE == "UP", 0.1, -0.1)
  )

line_data <- lysogeny_kegg %>%
  distinct(KEGG, DE)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = as.numeric(KEGG) + 0.1),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = as.numeric(KEGG) - 0.1),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 0.75, y = as.numeric(KEGG) + 0.1, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 0.75, y = as.numeric(KEGG) - 0.1, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = lysogeny_kegg,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      shape = Group
    )
  ) +
  scale_x_continuous(
    limits = c(0, 0.75),
    breaks = c(0, 0.25, 0.5, 0.75),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_y_continuous(
    breaks = seq_along(levels(lysogeny_kegg$KEGG)),
    labels = levels(lysogeny_kegg$KEGG)
  ) +
  scale_fill_viridis_c(limits = c(0, 15)) +
  scale_size_continuous(
    limits = c(1, 100),
    breaks = c(0, 10, 20, 30, 40)
  ) +
  scale_shape_manual(
    values = c(
      "Strains GIL01 vs GBJ002" = 21,
      "Strains GIL16 vs GBJ002" = 22,
      "Strains GIL16 vs GIL01" = 23
    )
  ) +
  labs(
    x = "Gene ratio",
    y = "KEGG subcategories",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

## DE genes

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_de_genes
#| code-fold: true

lysogeny_de_genes <- create_de_genes_table(group = "strains", filter = "Lysogeny")

colnames(lysogeny_de_genes) <- c("Gene ID", "Gene symbol", "Description", "GIL01 vs GBJ002", "GIL16 vs GBJ002", "GIL16 vs GIL01")

max_abs <- ceiling(max(abs(lysogeny_de_genes[, 4:ncol(lysogeny_de_genes)]), na.rm = TRUE))
brks <- seq(-max_abs, max_abs, length.out = 50)
clrs <- colorRampPalette(c("#5A8AC6", "white", "#F8696B"))(length(brks) + 1)

sketch <- withTags(table(
  class = "display",
  thead(
    tr(
      th(colspan = 3, style = "text-align: center; font-weight: bold;", "Gene annotation"),
      th(colspan = 3, style = "text-align: center; font-weight: bold;", "Lysogeny contrasts")
    ),
    tr(
      lapply(colnames(lysogeny_de_genes),
             function(x) th(style = "text-align: center;", x))
    )
  )
))

datatable(lysogeny_de_genes, container = sketch, rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 1),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = c(4:6), digits = 2) %>%
  formatStyle(columns = c(4:6), backgroundColor = styleInterval(brks, clrs)) %>%
  formatStyle(columns = c(1, 2, 4:ncol(lysogeny_de_genes)), `text-align` = "center")
```

## Enrichments summary

This table contains a summary of the enrichments for each category and contrast. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: lysogeny_es
#| code-fold: true

datatable(lysogeny_enrichments, rownames = FALSE, colnames = c("Group", "Contrast", "DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 3),
            scrollX = TRUE,
            scroller = TRUE,
            deferRender = TRUE))
```

# Intra-strain time-course summaries

This section presents the results of the intra-strain time-course contrasts. The two bar charts show the total number of DE genes and the number of enrichments in each contrast per strain. In addition, the enrichments in the COG and KEGG categories are represented in two bubble plots to facilitate comparison between contrasts. Finally, the interactive table compiles all the results from the **Enrichment summary** tables for each contrast.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_strain_time_course_summaries
#| code-fold: true

intra_strain_time_course_contrasts <- combine_summaries(type = "contrast", group = "time_points")
intra_strain_time_course_enrichments <- combine_summaries(type = "enrichment", group = "time_points")
```

## Barplots

This plot shows the number of DE genes, both upregulated and downregulated, in each contrast per strain. As we can see, there doesn't appear to be a difference in the number of DE genes according to the strain. This is because three independent models (i.e. three DESeq2 objects) were constructed for each set of contrasts, reflecting that the contrasts with fewer DE genes are those from early time points (0 and 10 minutes), while those performed with late time points (60 or 60-mock) show a large number of DE genes.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_strain_time_course_cs
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

intra_strain_time_course_contrasts_size <- intra_strain_time_course_contrasts %>%
  select(Group, Contrast, DE, Size) %>%
  mutate(
    Group = factor(Group, levels = levels(intra_strain_time_course_contrasts$Group)),
    Contrast = factor(Contrast, levels = levels(intra_strain_time_course_contrasts$Contrast)),
    DE = factor(DE, levels = levels(intra_strain_time_course_contrasts$DE))
  )

ggplot(intra_strain_time_course_contrasts_size, aes(x = Contrast, y = Size, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8),
           width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = Size),
    position = position_dodge(width = 0.8),
    vjust = -0.4,
    size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  facet_wrap(~ Group, ncol = 1) +
  ylim(0, 800) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold")) +
  labs(x = "Contrast", y = "DE genes", fill = "DE")
```

This plot shows the number of enrichments found in each category. Most enrichments are found in the BP GO category, unlike what happened in lysogeny, which were mainly found in MF and CC. In this plot, the contrasts between GBJ002 time points show less enrichments than with GIL01 or GIL16, which would reflect the differences found in gene expression throughout MMC induction (even without induction) when any of these tectiviruses are present.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_strain_time_course_ce
#| code-fold: true
#| warning: false
#| fig-width: 16
#| fig-height: 10

intra_strain_time_course_contrasts_enrichments <- intra_strain_time_course_contrasts %>%
  pivot_longer(
    cols = !c(Group, Contrast, DE, Size),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  filter(Category != "Size") %>%
  mutate(
    Group = factor(Group, levels = levels(intra_strain_time_course_contrasts$Group)),
    Contrast = factor(Contrast, levels = levels(intra_strain_time_course_contrasts$Contrast)),
    DE = factor(DE, levels = levels(intra_strain_time_course_contrasts$DE)),
    Category = factor(Category, levels = colnames(intra_strain_time_course_contrasts)[4:ncol(intra_strain_time_course_contrasts)][colnames(intra_strain_time_course_contrasts)[4:ncol(intra_strain_time_course_contrasts)] != "Size"])
  )

ggplot(intra_strain_time_course_contrasts_enrichments, aes(x = Category, y = Count, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = ifelse(Count == 0, NA, Count)),
    position = position_dodge(width = 0.8),
    vjust = -0.4, size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  scale_x_discrete(labels = c("COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)")) +
  facet_grid(Group ~ Contrast) +
  scale_y_continuous(limits = c(0, 70), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  ) +
  labs(x = NULL, y = "Enrichments", fill = "DE")
```

## Upset plots

Upset plots were used to evaluate the overlap of DE genes across the temporal contrasts analyzed within each strain (GBJ002, GIL01 and GIL16). Each set corresponds to the genes identified as significantly upregulated (in red) or downregulated (in blue) in a specific contrast and intersections represent genes shared among two or more contrasts. The size of each intersection reflects the number of genes exhibiting a common transcriptional response across the corresponding comparisons, whereas the set size indicates the total number of DE genes detected in each contrast. This analysis enables the identification of genes specifically associated with individual stages of the experiment as well as genes displaying persistent transcriptional changes across multiple time points, thereby providing insights into the dynamics and conservation of the transcriptional response within each strain.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: upset_intra
#| code-fold: true
#| warning: false
#| fig-width: 10
#| fig-height: 6

upset_contrasts(group = "GBJ002", de_genes = "UP", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))

upset_contrasts(group = "GBJ002", de_genes = "DOWN", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))

upset_contrasts(group = "GIL01", de_genes = "UP", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))

upset_contrasts(group = "GIL01", de_genes = "DOWN", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))

upset_contrasts(group = "GIL16", de_genes = "UP", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))

upset_contrasts(group = "GIL16", de_genes = "DOWN", order = c(
  "Time points 10 vs 0",
  "Time points 30 vs 0",
  "Time points 60 vs 0",
  "Time points 60 vs 60-mock",
  "Time points 60-mock vs 0"
))
```

## Bubble plots

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_cog
#| code-fold: true
#| fig-width: 14
#| fig-height: 6

intra_cog <- intra_cog %>%
  rename(Number = DE) %>%
  mutate(
    Group = Contrast,
    DE = Direction,
    COG_letter = str_extract(COG, "\\(([A-Z])\\)") %>% str_remove_all("[\\(\\)]"),
    minus_log10_p = -log10(p)
  )

category_order <- cog_names %>%
  filter(COG %in% intra_cog$COG_letter) %>%
  pull(COG)

ordered_categories <- intra_cog %>%
  mutate(COG_letter = factor(COG_letter, levels = category_order)) %>%
  arrange(desc(COG_letter)) %>%
  distinct(COG) %>%
  pull(COG)

intra_cog$COG <- factor(intra_cog$COG, levels = ordered_categories)
intra_cog$Group <- factor(intra_cog$Group, levels = c("Time points 10 vs 0", "Time points 30 vs 0", "Time points 60 vs 0", "Time points 60 vs 60-mock", "Time points 60-mock vs 0"))

intra_cog <- intra_cog %>%
  mutate(
    y_pos = as.numeric(COG) +
      ifelse(
        DE == "UP",
        runif(n(), 0.15, 0.35),
        runif(n(), -0.35, -0.15)
      )
  )

line_data <- intra_cog %>%
  distinct(COG, DE, Strain)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = as.numeric(COG) + 0.25),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = as.numeric(COG) - 0.25),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 1, y = as.numeric(COG) + 0.25, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 1, y = as.numeric(COG) - 0.25, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = intra_cog,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      shape = Group
    )
  ) +
  facet_wrap(~ Strain) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_y_continuous(
    breaks = seq_along(levels(intra_cog$COG)),
    labels = levels(intra_cog$COG)
  ) +
  scale_fill_viridis_c(limits = c(0, 10)) +
  scale_size_continuous(
    limits = c(1, 70),
    breaks = c(0, 10, 25, 50, 70)
  ) +
  scale_shape_manual(
    values = c(
      "Time points 10 vs 0"       = 21,
      "Time points 30 vs 0"       = 22,
      "Time points 60 vs 0"       = 23,
      "Time points 60 vs 60-mock" = 24,
      "Time points 60-mock vs 0"  = 25
    )
  ) +
  labs(
    x = "Gene ratio",
    y = "COG category",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_kegg
#| code-fold: true
#| fig-width: 14
#| fig-height: 6

intra_kegg <- intra_kegg %>%
  rename(Number = DE) %>%
  mutate(
    Group = Contrast,
    DE = Direction,
    minus_log10_p = -log10(p)
  )

category_order <- btn_pathways %>%
  pull(KEGG.pathway.maps)

ordered_categories <- intra_kegg %>%
  mutate(KEGG = factor(KEGG, levels = category_order)) %>%
  arrange(desc(KEGG)) %>%
  distinct(KEGG) %>%
  pull(KEGG)

intra_kegg$KEGG <- factor(intra_kegg$KEGG, levels = ordered_categories)
intra_kegg$Group <- factor(intra_kegg$Group, levels = c("Time points 10 vs 0", "Time points 30 vs 0", "Time points 60 vs 0", "Time points 60 vs 60-mock", "Time points 60-mock vs 0"))

intra_kegg <- intra_kegg %>%
  mutate(
    y_pos = as.numeric(KEGG) +
      ifelse(
        DE == "UP",
        runif(n(), 0.15, 0.35),
        runif(n(), -0.35, -0.15)
      )
  )

line_data <- intra_kegg %>%
  distinct(KEGG, DE, Strain)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = as.numeric(KEGG) + 0.25),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = as.numeric(KEGG) - 0.25),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 1, y = as.numeric(KEGG) + 0.25, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 1, y = as.numeric(KEGG) - 0.25, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = intra_kegg,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      shape = Group
    )
  ) +
  facet_wrap(~ Strain) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_y_continuous(
    breaks = seq_along(levels(intra_kegg$KEGG)),
    labels = levels(intra_kegg$KEGG)
  ) +
  scale_fill_viridis_c(limits = c(0, 10)) +
  scale_size_continuous(
    limits = c(1, 70),
    breaks = c(0, 10, 25, 50, 70)
  ) +
  scale_shape_manual(
    values = c(
      "Time points 10 vs 0"       = 21,
      "Time points 30 vs 0"       = 22,
      "Time points 60 vs 0"       = 23,
      "Time points 60 vs 60-mock" = 24,
      "Time points 60-mock vs 0"  = 25
    )
  ) +
  labs(
    x = "Gene ratio",
    y = "KEGG subcategories",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

## DE genes

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_de_genes
#| code-fold: true

intra_de_genes <- create_de_genes_table(group = "time_points")

colnames(intra_de_genes)[1:2] <- c("Gene ID", "Gene symbol")

max_abs <- ceiling(max(abs(intra_de_genes[, 4:ncol(intra_de_genes)]), na.rm = TRUE))
brks <- seq(-max_abs, max_abs, length.out = 50)
clrs <- colorRampPalette(c("#5A8AC6", "white", "#F8696B"))(length(brks) + 1)

lfc_cols <- colnames(intra_de_genes)[4:ncol(intra_de_genes)]
groups <- sub(" - .*", "", lfc_cols)
group_levels <- c("GBJ002", "GIL01", "GIL16")
group_counts <- sapply(group_levels, function(g) sum(groups == g))
second_row <- c(colnames(intra_de_genes)[1:3], sub(".* - ", "", lfc_cols))

sketch <- withTags(table(
  class = "display",
  thead(
    tr(
      th(colspan = 3, style = "text-align: center; font-weight: bold;", "Gene annotation"),
      lapply(seq_along(group_levels), function(i)
        th(colspan = group_counts[i], style = "text-align: center; font-weight: bold;", group_levels[i])
      )
    ),
    tr(
      lapply(second_row, function(x) th(style = "text-align: center;", x))
    )
  )
))

datatable(intra_de_genes, container = sketch, rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 1),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = 4:ncol(intra_de_genes), digits = 2) %>%
  formatStyle(columns = 4:ncol(intra_de_genes), backgroundColor = styleInterval(brks, clrs)) %>%
  formatStyle(columns = c(1, 2, 4:ncol(intra_de_genes)), `text-align` = "center")
```

## Enrichments summary

This table contains a summary of the enrichments for each category, contrast and strain. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: intra_strain_time_course_es
#| code-fold: true

datatable(intra_strain_time_course_enrichments, rownames = FALSE, colnames = c("Strain", "Contrast", "DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 3),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE))
```

# Inter-strain time-course summaries

This section presents the results of the inter-strain time-course contrasts. The two bar charts show the total number of DE genes and the number of enrichments in each contrast per strain. In addition, the enrichments in the COG and KEGG categories are represented in two bubble plots to facilitate comparison between contrasts. Finally, the interactive table compiles all the results from the **Enrichment summary** tables for each contrast.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_strain_time_course_summaries
#| code-fold: true

inter_strain_time_course_contrasts <- combine_summaries(type = "contrast", group = "strains", filter = c("t0", "t10", "t30", "t60", "t60-mock"))
inter_strain_time_course_enrichments <- combine_summaries(type = "enrichment", group = "strains", filter = c("t0", "t10", "t30", "t60", "t60-mock"))
```

## Barplots

This plot shows the number of DE genes, both upregulated and downregulated, in each time per strains contrasts. As we can see, the highest number of DE genes are obtained at late time points, and especially in the contrasts of GBJ002 vs GIL01 and GIL16. This is because a combined model was used for all contrasts (i.e. a single DESeq2 object), reflecting that, considering both time and strain, the greatest transcriptional changes are found at late time points.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_strain_time_course_cs
#| code-fold: true
#| fig-width: 10
#| fig-height: 8

inter_strain_time_course_contrasts_size <- inter_strain_time_course_contrasts %>%
  select(Group, Contrast, DE, Size) %>%
  mutate(
    Group = factor(Group, levels = levels(inter_strain_time_course_contrasts$Group)),
    Contrast = factor(Contrast, levels = levels(inter_strain_time_course_contrasts$Contrast)),
    DE = factor(DE, levels = levels(inter_strain_time_course_contrasts$DE))
  )

ggplot(inter_strain_time_course_contrasts_size, aes(x = Group, y = Size, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8),
           width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = Size),
    position = position_dodge(width = 0.8),
    vjust = -0.4,
    size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  scale_x_discrete(labels = c(0, 10, 30, 60, "60-mock")) +
  facet_wrap(~ Contrast, ncol = 1) +
  ylim(0, 700) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold")) +
  labs(x = "Time points (min)", y = "DE genes", fill = "DE")
```

This plot shows the number of enrichments found in each category. Most enrichments, as in the previous plot, are found at late time points in the contrasts of GBJ002 vs GIL01 and GIL16. In contrast, the differences between both strains with tectivirus, although they exist, are minimal and are only found at the 30 and 60-mock time points.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_strain_time_course_ce
#| code-fold: true
#| warning: false
#| fig-width: 16
#| fig-height: 10

inter_strain_time_course_contrasts_enrichments <- inter_strain_time_course_contrasts %>%
  pivot_longer(
    cols = !c(Group, Contrast, DE, Size),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  filter(Category != "Size") %>%
  mutate(
    Group = factor(Group, levels = levels(inter_strain_time_course_contrasts$Group)),
    Contrast = factor(Contrast, levels = levels(inter_strain_time_course_contrasts$Contrast)),
    DE = factor(DE, levels = levels(inter_strain_time_course_contrasts$DE)),
    Category = factor(Category, levels = colnames(inter_strain_time_course_contrasts)[4:ncol(inter_strain_time_course_contrasts)][colnames(inter_strain_time_course_contrasts)[4:ncol(inter_strain_time_course_contrasts)] != "Size"])
  )

ggplot(inter_strain_time_course_contrasts_enrichments, aes(x = Category, y = Count, fill = DE)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, alpha = 0.7) +
  geom_text(
    aes(label = ifelse(Count == 0, NA, Count)),
    position = position_dodge(width = 0.8),
    vjust = -0.4, size = 4
  ) +
  scale_fill_manual(values = c("UP" = "red", "DOWN" = "blue")) +
  scale_x_discrete(labels = c("COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)")) +
  facet_grid(Contrast ~ Group,
    labeller = labeller(Group =
      c("t0" = "Time point 0",
        "t10" = "Time point 10",
        "t30" = "Time point 30",
        "t60" = "Time point 60",
        "t60-mock" = "Time point 60-mock"))) +
  scale_y_continuous(limits = c(0, 60), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  ) +
  labs(x = NULL, y = "Enrichments", fill = "DE")
```

## Upset plots

Upset plots were used to examine the overlap of DE genes among pairwise strain comparisons performed at different time points (30, 60 and 60-mock). Separate analyses were conducted for upregulated (in red) and downregulated (in blue) genes. Each set represents the genes identified in a specific strain comparison at a given time point, whereas intersections correspond to genes shared across multiple pairwise comparisons. The size of each intersection indicates the number of genes exhibiting consistent differential expression patterns in more than one strain comparison, while the set size reflects the total number of DE genes detected in each individual contrast. This approach enables the identification of genes contributing to common transcriptional differences among strains, as well as genes specifically associated with particular strain comparisons or time points.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: upset_inter
#| code-fold: true
#| warning: false
#| fig-width: 10
#| fig-height: 6

upset_contrasts(group = c("t30", "t60", "t60-mock"), de_genes = "UP", order = c(
  "Time point 30 - Strains GIL01 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GIL01",
  "Time point 60 - Strains GIL01 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GIL01",
  "Time point 60-mock - Strains GIL01 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GIL01"
))

upset_contrasts(group = c("t30", "t60", "t60-mock"), de_genes = "DOWN", order = c(
  "Time point 30 - Strains GIL01 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GIL01",
  "Time point 60 - Strains GIL01 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GIL01",
  "Time point 60-mock - Strains GIL01 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GIL01"
))
```

## Bubble plots

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_cog
#| code-fold: true
#| fig-width: 10
#| fig-height: 6

inter_cog <- inter_cog %>%
  rename(Number = DE) %>%
  mutate(
    DE = Direction,
    Group = paste0("Time point ", Time_point, " - ", Contrast),
    COG_letter = str_extract(COG, "\\(([A-Z])\\)") %>% str_remove_all("[\\(\\)]"),
    minus_log10_p = -log10(p)
  )

category_order <- cog_names %>%
  filter(COG %in% inter_cog$COG_letter) %>%
  pull(COG)

ordered_categories <- inter_cog %>%
  mutate(COG_letter = factor(COG_letter, levels = category_order)) %>%
  arrange(desc(COG_letter)) %>%
  distinct(COG) %>%
  pull(COG)

inter_cog$COG <- factor(inter_cog$COG, levels = ordered_categories)
inter_cog$Group <- factor(inter_cog$Group, levels = c(
  "Time point 30 - Strains GIL16 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GIL01",
  "Time point 60 - Strains GIL01 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GIL01"))

inter_cog <- inter_cog %>%
  mutate(
    y_pos = as.numeric(COG) + ifelse(DE == "UP", 0.1, -0.1)
  )

line_data <- inter_cog %>%
  distinct(COG, DE, y_pos)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = y_pos),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = y_pos),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 0.2, y = y_pos, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 0.2, y = y_pos, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = inter_cog,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      colour = minus_log10_p,
      shape = Group
    )
  ) +
  scale_x_continuous(limits = c(0, 0.2), expand = expansion(mult = c(0, 0.15))) +
  scale_y_continuous(
    breaks = seq_along(levels(inter_cog$COG)),
    labels = levels(inter_cog$COG)
  ) +
  scale_fill_viridis_c(limits = c(0, 20)) +
  scale_colour_viridis_c(limits = c(0, 20), guide = "none") +
  scale_size_continuous(limits = c(1, 50), breaks = c(0, 10, 20, 30, 40, 50)) +
  scale_shape_manual(values = c(21, 12, 22, 23, 24, 25)) +
  labs(
    x = "Gene ratio",
    y = "COG category",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_kegg
#| code-fold: true
#| fig-width: 10
#| fig-height: 6

inter_kegg <- inter_kegg %>%
  rename(Number = DE) %>%
  mutate(
    DE = Direction,
    Group = paste0("Time point ", Time_point, " - ", Contrast),
    minus_log10_p = -log10(p)
  )

category_order <- btn_pathways %>%
  pull(KEGG.pathway.maps)

ordered_categories <- inter_kegg %>%
  mutate(KEGG = factor(KEGG, levels = category_order)) %>%
  arrange(desc(KEGG)) %>%
  distinct(KEGG) %>%
  pull(KEGG)

inter_kegg$KEGG <- factor(inter_kegg$KEGG, levels = ordered_categories)
inter_kegg$Group <- factor(inter_kegg$Group, levels = c(
  "Time point 10 - Strains GIL16 vs GBJ002",
  "Time point 30 - Strains GIL01 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GBJ002",
  "Time point 30 - Strains GIL16 vs GIL01",
  "Time point 60 - Strains GIL01 vs GBJ002",
  "Time point 60 - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GBJ002",
  "Time point 60-mock - Strains GIL16 vs GIL01"))

inter_kegg <- inter_kegg %>%
  mutate(
    y_pos = as.numeric(KEGG) + ifelse(DE == "UP", 0.1, -0.1)
  )

line_data <- inter_kegg %>%
  distinct(KEGG, DE, y_pos)

ggplot() +
  geom_hline(
    data = subset(line_data, DE == "UP"),
    aes(yintercept = y_pos),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_hline(
    data = subset(line_data, DE == "DOWN"),
    aes(yintercept = y_pos),
    colour = "blue",
    linetype = "dashed",
    linewidth = 0.75,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "UP"),
    aes(x = 0.6, y = y_pos, label = "UP"),
    colour = "red",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_text(
    data = subset(line_data, DE == "DOWN"),
    aes(x = 0.6, y = y_pos, label = "DOWN"),
    colour = "blue",
    vjust = -0.25,
    fontface = "bold",
    size = 3,
    alpha = 0.5
  ) +
  geom_point(
    data = inter_kegg,
    aes(
      x = Ratio,
      y = y_pos,
      size = Number,
      fill = minus_log10_p,
      colour = minus_log10_p,
      shape = Group
    )
  ) +
  scale_x_continuous(limits = c(0, 0.6), expand = expansion(mult = c(0, 0.15))) +
  scale_y_continuous(
    breaks = seq_along(levels(inter_kegg$KEGG)),
    labels = levels(inter_kegg$KEGG)
  ) +
  scale_fill_viridis_c(limits = c(0, 31)) +
  scale_colour_viridis_c(limits = c(0, 31), guide = "none") +
  scale_size_continuous(limits = c(1, 50), breaks = c(0, 5, 10, 20, 30)) +
  scale_shape_manual(values = c(4, 21, 22, 23, 24, 8, 25, 10)) +
  labs(
    x = "Gene ratio",
    y = "KEGG subcategories",
    size = "DE genes",
    fill = expression(paste("Adjusted p value (", -log[10], ")")),
    shape = "Contrast"
  ) +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank()
  )
```

## DE genes

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_de_genes
#| code-fold: true
#| warning: false

inter_de_genes <- create_de_genes_table(group = "strains", filter = c("t0", "t30", "t60", "t60-mock"))

colnames(inter_de_genes)[1:2] <- c("Gene ID", "Gene symbol")

max_abs <- ceiling(max(abs(inter_de_genes[, 4:ncol(inter_de_genes)]), na.rm = TRUE))
brks <- seq(-max_abs, max_abs, length.out = 50)
clrs <- colorRampPalette(c("#5A8AC6", "white", "#F8696B"))(length(brks) + 1)

lfc_cols <- colnames(inter_de_genes)[4:ncol(inter_de_genes)]
groups <- sub(" - .*", "", lfc_cols)
group_levels <- c("t0", "t30", "t60", "t60-mock")
group_labels <- c("Time point 0", "Time point 30", "Time point 60", "Time point 60-mock")
group_counts <- sapply(group_levels, function(g) sum(groups == g))
second_row <- c(colnames(inter_de_genes)[1:3], sub(".* - ", "", lfc_cols))

sketch <- withTags(table(
  class = "display",
  thead(
    tr(
      th(colspan = 3, style = "text-align: center; font-weight: bold;", "Gene annotation"),
      lapply(seq_along(group_levels), function(i)
        th(colspan = group_counts[i], style = "text-align: center; font-weight: bold;", group_labels[i])
      )
    ),
    tr(
      lapply(second_row, function(x) th(style = "text-align: center;", x))
    )
  )
))

datatable(inter_de_genes, container = sketch, rownames = FALSE,
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 1),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE)) %>%
  formatRound(columns = 4:ncol(inter_de_genes), digits = 2) %>%
  formatStyle(columns = 4:ncol(inter_de_genes), backgroundColor = styleInterval(brks, clrs)) %>%
  formatStyle(columns = c(1, 2, 4:ncol(inter_de_genes)), `text-align` = "center")

counts <- lysogeny_counts
tectivirus_de_table <- create_de_tectivirus_genes_table()

colnames(tectivirus_de_table) <- c("Gene symbol", "Description", "Lysogeny", "Time point 0", "Time point 10", "Time point 30", "Time point 60", "Time point 60-mock")

max_abs <- ceiling(
  max(
    abs(tectivirus_de_table[, 3:ncol(tectivirus_de_table)]),
    na.rm = TRUE
  )
)

brks <- seq(-max_abs, max_abs, length.out = 50)

clrs <- colorRampPalette(
  c("#5A8AC6", "white", "#F8696B")
)(length(brks) + 1)

sketch <- withTags(
  table(
    class = "display",
    thead(
      tr(
        th(
          colspan = 2,
          style = "text-align: center; font-weight: bold;",
          "Gene annotation"
        ),
        th(
          colspan = 6,
          style = "text-align: center; font-weight: bold;",
          "Strains GIL16 vs GIL01 contrasts"
        )
      ),
      tr(
        lapply(
          colnames(tectivirus_de_table),
          function(x) th(style = "text-align: center;", x)
        )
      )
    )
  )
)

datatable(
  tectivirus_de_table,
  container = sketch,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns", "Scroller"),
  options = list(
    dom = "Brftip",
    buttons = c("copy", "csv", "excel"),
    fixedColumns = list(leftColumns = 1),
    scrollX = TRUE,
    scrollY = 400,
    scroller = TRUE,
    deferRender = TRUE
  )
) %>%
  formatRound(columns = 3:ncol(tectivirus_de_table), digits = 2) %>%
  formatStyle(
    columns = 3:ncol(tectivirus_de_table),
    backgroundColor = styleInterval(brks, clrs)
  ) %>%
  formatStyle(
    columns = c(1, 3:ncol(tectivirus_de_table)),
    `text-align` = "center"
  )
```

## Enrichments summary

This table contains a summary of the enrichments for each category, contrast and time. The subcategory was used for KEGG and the description for GO. Although you can scroll, I recommend exporting using the Excel button and viewing the table in full screen.

```{r}
#| echo: true
#| eval: true
#| error: true
#| label: inter_strain_time_course_es
#| code-fold: true

datatable(inter_strain_time_course_enrichments, rownames = FALSE, colnames = c("Time point", "Contrast", "DE genes", "DE number", "COG", "KEGG", "GO (BP)", "GO (MF)", "GO (CC)", "EC (1)", "EC (2)", "EC (3)"),
          filter = "top", extensions = c("Buttons", "FixedColumns", "Scroller"), options = list(
            autoWidth = TRUE,
            dom = "Brftip",
            buttons = c("copy", "csv", "excel"),
            fixedColumns = list(leftColumns = 3),
            scrollX = TRUE,
            scrollY = 400,
            scroller = TRUE,
            deferRender = TRUE))
```